### 2025-04-16 Examples/amspaperV5.tex
Initial Overleaf Import

%% Version 5.0, 2 January 2020
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% amspaperV5.tex --  LaTeX-based template for submissions to American 
% Meteorological Society journals
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Start with one of the following:
% DOUBLE-SPACED VERSION FOR SUBMISSION TO THE AMS
\documentclass{ametsocV5}

% TWO-COLUMN JOURNAL PAGE LAYOUT---FOR AUTHOR USE ONLY
% \documentclass[twocol]{ametsocV5}

% Enter packages here. If too many math alphabets are used,
% remove unnecessary packages or define hmmax and bmmax as necessary.

%\newcommand{\hmmax}{0}
%\newcommand{\bmmax}{0}
\usepackage{amsmath,amsfonts,amssymb,bm}
\usepackage{mathptmx}%{times}
\usepackage{newtxtext}
\usepackage{newtxmath}

%%% To be entered by author:

%% May use \\ to break lines in title:

\title{Title here}

%%% Enter authors' names, as you see in this example:
%%% Use \correspondingauthor{} and \thanks{Current Affiliation:...}
%%% immediately following the appropriate author.
%%%
%%% Note that the \correspondingauthor{} command is NECESSARY.
%%% The \thanks{} commands are OPTIONAL.

    \authors{Author One\correspondingauthor{Author name, email address}
 and Author Two\thanks{Current affiliation: NCAR, Boulder, Colorado.}}

     \affiliation{American Meteorological Society, 
     Boston, Massachusetts}

    \extraauthor{Extra Author}
    \extraaffil{Affiliation, City, State/Province, Country}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%
% Enter your Abstract here

\abstract{Enter the text of your abstract here.  This is a sample American
Meteorological Society (AMS) \LaTeX\ template.  This document provides
authors with instructions on the use of the AMS \LaTeX\ template.  Authors
should refer to the file amspaper.tex to review the actual \LaTeX\ code used
to create this document. The template.tex
file should be modified by authors for their own manuscript.}

\begin{document}

%% Necessary!
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SIGNIFICANCE STATEMENT/CAPSULE SUMMARY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% If you are including an optional significance statement for a journal article or a required capsule summary for BAMS 
% (see www.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/formatting-and-manuscript-components for details), 
% please apply the necessary command as shown below:
%
% \statement
% Significance statement here.
%
% \capsule
% Capsule summary here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAIN BODY OF PAPER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

This document will provide authors with the basic American Meteorological
Society (AMS) formatting guidelines. This document was created using \LaTeX\
and demonstrates how to use the \LaTeX\ template when submitting a manuscript
to the AMS.  The following sections will outline the guidelines and
formatting for text, math, figures, and tables while using \LaTeX\/ for a
submission to the AMS. An attempt to compile amspaper.tex should be made
before using the template. The files have been tested on Windows, Linux, and
Mac OS using \TeX\ Live 2019 (available online at
\url{http://www.tug.org/texlive/}). Feedback and questions should be sent to
latex@ametsoc.org. Additional information is available on the AMS \LaTeX\ Submission Info
web page (\url{http://www2.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/author-resources/latex-author-info/}).

Authors should use the empty template.tex to begin their
paper. A valuable source of \LaTeX\ information is the {TeX
Frequently Asked Questions} page (available online at \url{faq.tug.org}).

\section{Formatting text and sections}
The text should be divided into sections, each with a separate heading and
consecutive numbering. Note, however, that single secondary, tertiary, and
quaternary sections remain unnumbered. Each section heading should be placed
on a separate line using the appropriate \LaTeX\ commands.

\subsection*{Secondary headings} 
Secondary headings labeled with letters are formatted using the
$\backslash$subsection$*${\tt\string{\string}} for a single
subsection within a section or
or $\backslash$subsection{\tt\string{\string}} 
for multiple
subsections within one section.

\subsubsection*{Tertiary headings} 
Tertiary headings are formatted using the
$\backslash$subsubsection$*${\tt\string{\string}} for single a subsubsection
within a subsection or
$\backslash$subsubsection{\tt\string{\string}} 
for multiple subsubsections
within a subsection.

\paragraph*{Quaternary headings} 
Quaternary headings are formatted using the
$\backslash$paragraph$*${\tt\string{\string}} for a single paragraph within
a subsubsection or
$\backslash$paragraph{\tt\string{\string}} 
for multiple paragraphs
within a subsection.

\section{Citations}
Citations to standard references in text should consist of the name of the
author and the year of publication, for example, \citet{Becker+Schmitz2003} or
\citep{Becker+Schmitz2003} using the appropriate $\backslash$citet\ or
$\backslash$citep commands, respectively. A variety of citation formats can
be used with the natbib package; however, the AMS prefers that authors use only the $\backslash$citet\ and
$\backslash$citep commands. References should be entered in the references.bib file. For a thorough
discussion of how to enter references into the references.bib database file
following AMS style, please refer to the \textbf{AMS\_RefsV5.pdf} document
included in this package.

\section{Formatting math}
The following sections will outline the basic formatting rules for
mathematical symbols and units.  In addition, a review of the amspaper.tex
file will show how this is done with the use of \LaTeX\ commands.  The AMS
template provides the American Mathematical Society math, font, symbol, and
boldface packages for use in math mode.

\subsection{Mathematical symbols}
Symbols must be of the same font style both in text discussion and in
displayed equations or terms (and figures should be prepared to match).
Scalar single-character symbols are set italic, Greek, or script. Examples
are $u$, $L$ [note that $\upsilon$ (Greek upsilon) is used instead of
\textit{v} (italic ``vee'') to avoid confusion with $\nu$ (Greek nu) often
used for viscosity; this is handled automatically when in \LaTeX\ math mode],
$w$, $x$, $y$, $z$, $f$, $g$, $r$, indices such as $i$ or $j$, and constants
such as $C_D$, $k$, or $K$. Multiple-character scalar variables,
abbreviations, nondimensional numbers, and acronyms for variables are set
regular nonitalic: $\mathrm{LWC}$, $\mathrm{Re}$, $\mathrm{Ro}$,
$\mathrm{BT}$, $\mathrm{abs}$, $\mathrm{obs}$, $\mathrm{max}$,
$\mathrm{min}$, $\mathrm{Re}$/$\mathrm{Im}$ (real/imaginary), etc. For
vectors, use boldface nonitalic Times Roman as in $\mathbf{V}$, $\mathbf{v}$,
or $\mathbf{x}$, and $\mathbf{i}$, $\mathbf{j}$, and $\mathbf{k}$ unit
vectors. Do not use the \LaTeX\ $\backslash$vec command to denote vectors.
For matrix notation, use nonitalic boldface Arial (or sans serif) font as in
$\bm{\mathsf{A}}$, $\bm{\mathsf{B}}$, or $\bm{\mathsf{M}}$.  All mathematical
operator abbreviations/acronyms are set lowercase regular Roman font, except
$O$ (on the order of): $\sin$, $\cos$, $\tan$, $\tanh$, $\mathrm{cov}$, $\Pr$
(for probability; note same as Prandtl number), $\mathrm{const}$ (for
constant), $\mathrm{c.c.}$ (complex conjugate).

\subsection{Units}
Units are always set on a single line with a space separating the
denominator, which is set with a superscript $-1$, $-2$, and so on, rather
than using a slash for ``per.'' Examples are g kg$^{-1}$, m$^2$ s$^{-1}$, W
m$^{-2}$, g m$^{-3}$, and m s$^{-1}$ (note that ms$^{-1}$ is the unit for
``per millisecond'').

\subsection{Equations}
Brief equations or terms set inline in text must be set as a single-line
expression because page proofs are not double spaced, for example,
$\rho^{-1}p/x$ or $(1/{\rho})p/x$ or $(a-b)/(c+d)$; that is, use a
superscript $-1$ for the denominator. In case of a more complicated term or
equation, it should be set as an unnumbered display equation, such as
\begin{displaymath} x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  \end{displaymath}

Otherwise, numbered display equations can be entered using the appropriate
equation command, such as \begin{equation}
x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  \end{equation}

Lists of equations are punctuated as written English, and commas, semicolons,
and periods are placed where appropriate. Conjunctions such as ``and,''
``while,'' ``when,'' or ``for'' are also typically placed before the final
element in a mathematical phrase, as befits the intended mathematical
meaning.

\section{Figures and tables}
The AMS prefers that all figures and tables are placed {\bf at the end of the document} prior to submission. 
A list of tables and a list of figures will appear near the end of the PDF
file, before the actual tables and figures. These lists are necessary for submission.

For appendix figures and tables, special commands are needed to manually 
change the numbering to ensure that each appendix figure or table is numbered 
as part of the respective appendix and not as a continuation of the main paper. Use the command
$\backslash$appendcaption\{\} instead of the usual $\backslash$caption\{\} to adjust the 
numbering; for example, for Table A1, you would use the command \verb+\appendcaption{A1}+.

Note that the normal $\backslash$ref\{\} command cannot be used to cite 
appendix figures and tables as the numbering will be incorrect. Callouts for appendix figures 
and tables in the text will need to be written out as plain text, for example, Fig. A1 and Table A1.

\subsection{Figures}
The insertion of a sample figure (Fig. \ref{f1})  
and caption is given below (in the .tex document) and at the end of the document. Standard figure sizes are 19 (one column), 
27, 33, and 39 (two columns) picas.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIGURES---PLACE AT END OF DOCUMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[h]
% \centerline{\includegraphics[width=19pc]{figure01.pdf}}
%  \caption{Enter the caption for your figure here.  Repeat as
%  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
%\end{figure}
%

\subsection{Tables}
Each table must be numbered, provided with a caption, and mentioned
specifically in the text. 
See below (in the .tex document) and at the end of the document for the formatting of a sample table (Table \ref{t1}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLES---PLACE AT END OF DOCUMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{table}[h]
%\caption{This is a sample table caption and table layout.  
%Table from Lorenz (1963).}\label{t1}
%\begin{center}
%\begin{tabular}{ccccrrcrc}
%\topline
%$N$ & $X$ & $Y$ & $Z$\\
%\midline
% 0000 & 0000 & 0010 & 0000 \\
% 0005 & 0004 & 0012 & 0000 \\
% 0010 & 0009 & 0020 & 0000 \\
% 0015 & 0016 & 0036 & 0002 \\
% 0020 & 0030 & 0066 & 0007 \\
% 0025 & 0054 & 0115 & 0024 \\
%\botline
%\end{tabular}
%\end{center}
%\end{table}
%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments
Keep acknowledgments (note correct spelling: no ``e'' between the ``g'' and
``m'') as brief as possible. In general, acknowledge only direct help in
writing or research. Financial support (e.g., grant numbers) for the work
done, for an author, or for the laboratory where the work was performed is
best acknowledged here rather than as footnotes to the title or to an
author's name. Contribution numbers (if the work has been published by the
author's institution or organization) should be included as footnotes on the title page,
not in the acknowledgments.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DATA AVAILABILITY STATEMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
%
\datastatement
The data availability statement is where authors should describe how the data underlying 
the findings within the article can be accessed and reused. Authors should attempt to 
provide unrestricted access to all data and materials underlying reported findings. 
If data access is restricted, authors must mention this in the statement.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIXES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% If only one appendix, use
%%\appendix%

%% If more than one appendix, use \appendix[<letter>], e.g.,
 \appendix[A]

\appendixtitle{Title of Appendix}

\subsection*{Appendix section}

The AMS template allows authors to format an unlimited number of
appendixes. To format a single appendix, use the $\backslash$appendix
command with no additional argument. Otherwise, add the appropriate
one-letter argument to the $\backslash$appendix command (e.g.
$\backslash$appendix[A], $\backslash$appendix[B],
$\backslash$appendix[C], etc.) corresponding to the appropriate
appendix.

The title of the appendix can be formatted using the
$\backslash$appendixtitle{\tt\string{\string}} 
 command. The $\backslash$subsection, $\backslash$subsubection,
and $\backslash$paragraph commands are used to create sections within
the appendix. (Note that the appendix title takes the place of $\backslash$section 
in the appendix, so the first section should begin with $\backslash$subsection
instead of $\backslash$section.)
 Equations are automatically numbered appropriately for 
each appendix. Here is an example of the first equation in appendix
A, automatically labeled (\ref{eq:1}): 
\begin{equation} \label{eq:1}
x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
\end{equation}

For appendix figures and tables, special commands are needed to manually 
change the numbering to ensure that each appendix figure or table is numbered 
as part of the appendix and not as a continuation of the main paper. Use the command
$\backslash$appendcaption\{\} instead of the usual $\backslash$caption\{\} to adjust the 
numbering; for example, for Table A1, you would use the command \verb+\appendcaption{A1}+.
In-text callouts for each appendix figure and table will need to be written as plain text;
the usual $\backslash$ref\{\} command cannot be used.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%APPENDIX FIGURE AND TABLE EXAMPLES---PLACE AT END OF DOCUMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\begin{table}
%\appendcaption{A1}{Here is the appendix table caption.}
%\centering
%\begin{tabular}{ccc}
%a&b&c\\
%d&e&f
%\end{tabular}
%\end{table}
%
%\begin{figure}
%\centerline{(illustration here)}
%\appendcaption{A1}{Here is the appendix figure caption.}
%\end{figure}
%

\appendix[B]
\appendixtitle{File Structure of the AMS \LaTeX\ Package}

\subsection{AMS \LaTeX\ files}
You will be provided with a tarred, zipped \LaTeX\ package containing 
17 files. These files are

\begin{description}
\item[Basic style file:] ametsocV5.cls.

The file ametsocV5.cls is the manuscript style file.

\begin{itemize}
\item
Using \verb+\documentclass{ametsocV5}+ for your .tex document
will 
generate a PDF that follows all AMS guidelines for submission and peer
review.

\item
Using \verb+\documentclass[twocol]{ametsocV5}+ for your .tex document
can be used to generate a PDF that closely
follows the layout of an AMS journal page, including single spacing and two
columns.  This journal style PDF is only for the author's personal use, and
any papers submitted in this style will not be accepted.  
\end{itemize}
Always use \verb+\documentclass{ametsocV5}+ 
when generating a PDF for submission to the AMS.

\item[Template:]
templateV5.tex, for the author to use when making his/her
paper.
The file provides a basic blank template with some
section headings for authors to easily enter their manuscript.

\item[Sample .tex and .pdf files:]
The file amspapeV5r.tex contains the \LaTeX\ code for the sample file.  
The resulting PDF can be seen in amspaper.pdf (this file).

\item[Sample article:] article formatted in draft and two-column mode.

\begin{itemize}
\item
AMSSamp1V5.tex, AMSSamp1V5.pdf\\
Formal paper done in draft mode and the resulting .pdf.

\item
AMSSamp2V5.tex, AMSSamp2V5.pdf \\
The same paper using the \verb+[twocol]+ option and the resulting .pdf.

\item
FigOne.pdf, FigTwo.pdf, and figure01.pdf are sample figures.

\end{itemize}

\item[Bibliography Files:]

ametsoc2014.bst, database2020.bib, and references.bib.

\begin{itemize}
\item
ametsoc2014.bst is the bibliography style file.

\item
database2020.bib is an example of a bibliographic database file.

\item
references.bib should be altered with your own bibliography information.  
\end{itemize}

\item[Documention:] found in AMSDocsV5.pdf. Additional information
found in
readme.txt, which contains a list of the files and how they are used.

\end{description}

\subsection{Help for Authors}
Questions and feedback concerning the use of the AMS \LaTeX\ files should be
directed to latex@ametsoc.org. Additional information is available on the AMS
\LaTeX\ Submission Info web page (\url{http://www2.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/author-resources/latex-author-info/}).

\appendix[C]
\appendixtitle{Building a PDF and Submitting Your 
\LaTeX\ Manuscript Files to the AMS}

\subsection{Building your own PDF}
There are a variety of different methods and programs that will create a
final PDF from your \LaTeX\ files. The easiest method is to download one of
the freely available text editors/compilers such as TexWorks or TeXnicCenter.
TexWorks is installed with the TeXLive distribution and provides both a text
editor and the ability to compile your files into a PDF.

\subsection{Submitting your files to the AMS for peer review}
The AMS uses the Editorial Manager system for all author submissions for
peer review. Editorial Manager uses the freely available \TeX\ Live 2018
distribution. This system will automatically generate a PDF from your
submitted \LaTeX\ files and figures.

You should not upload your own PDF into
the system. If the system does not build the PDF from your files correctly,
refer to the AMS \LaTeX\ FAQ page first for possible solutions. If your PDF
still does not build correctly after trying the solutions on the FAQ page, email
latex@ametsoc.org for help.

\subsection{Other software}
As mentioned above, there is a variety of software that can be used to edit
.tex files and build a PDF.  The AMS does not support \LaTeX\/-related
WYSIWYG software, such as Scientific Workplace, or WYSIWYM software, such as
LyX.  \TeX\ Live (available online at \\ \url{http://www.tug.org/texlive/}) is
recommended for users needing an up-to-date \LaTeX\ distribution with
software that includes an editor and the ability to automatically generate a
PDF.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 This shows how to enter the commands for making a bibliography using
 BibTeX. It uses references.bib and the ametsoc2014.bst file for the style.

\bibliographystyle{ametsoc2014}
\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[h]
\caption{This is a sample table caption and table layout.}\label{t1}
\begin{center}
\begin{tabular}{ccccrrcrc}
\topline
$N$ & $X$ & $Y$ & $Z$\\
\midline
 0000 & 0000 & 0010 & 0000 \\
 0005 & 0004 & 0012 & 0000 \\
 0010 & 0009 & 0020 & 0000 \\
 0015 & 0016 & 0036 & 0002 \\
 0020 & 0030 & 0066 & 0007 \\
 0025 & 0054 & 0115 & 0024 \\
\botline
\end{tabular}
\end{center}
\end{table}
%

\begin{table}
\appendcaption{A1}{Here is the appendix table caption.}
\centering
\begin{tabular}{ccc}
\topline
$1$ & $2$ & $3$ \\
\midline
a&b&c \\
d&e&f \\
\botline
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIGURES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
 \centerline{\includegraphics[width=19pc]{figure01.pdf}}
  \caption{Enter the caption for your figure here.  Repeat as
  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
\end{figure}

\begin{figure}
\centerline{(illustration here)}
\appendcaption{A1}{Here is the appendix figure caption.}
\end{figure}

\begin{figure}
\centerline{(illustration here)}
\appendcaption{B1}{Here is the appendix figure caption.}
\end{figure}

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% END OF AMSPAPER.TEX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-04-16 Examples/amssamp1V5.tex
Initial Overleaf Import

%% amssamp1V5.tex is nearly identical to amssamp2.tex, except
%% that amssamp2V5.tex uses the [twocol] option to produce
%% two-column text.

\documentclass{ametsocV5}

%\documentclass[twocol]{ametsocV5}

% Enter packages here. If too many math alphabets are used,
% remove unnecessary packages or define hmmax and bmmax as necessary.

%\newcommand{\hmmax}{0}
%\newcommand{\bmmax}{0}
\usepackage{amsmath,amsfonts,amssymb,bm}
\usepackage{mathptmx}%{times}
\usepackage{newtxtext}
\usepackage{newtxmath}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Citations should be of the form ``author year''  not ``author, year''
%\bibpunct{(}{)}{;}{a}{}{,}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Dominant Anomaly Patterns in the Near-Surface 
Baroclinicity and Accompanying
Anomalies in the Atmosphere and Oceans. 
Part II: North Pacific Basin}

\authors{Mototaka Nakamura\correspondingauthor{Mototaka Nakamura, moto@jamstec.go.jp} and Shozo Yamane\thanks{Current affiliation: 
Science and Engineering, Doshisha University, Kyotanabe, Kyoto, Japan.}}

\affiliation{Japan Agency for Marine-Earth Science and Technology, Yokohama,
Kanagawa, Japan}

%\extraauthor{Extra Author}
%\extraaffil{Affiliation, City, State/Province, Country}

\abstract{Variability in the monthly-mean flow and storm track in the North
Pacific basin is examined with a focus on the near-surface baroclinicity.
Dominant patterns of anomalous near-surface baroclinicity found from
EOF analyses generally show mixed patterns of
shift and changes in the strength of near-surface baroclinicity. Composited
anomalies in the monthly-mean wind at various pressure levels based on the
signals in the EOFs show accompanying anomalies in the mean flow up to 50 hPa
in the winter and up to 100 hPa in other seasons. Anomalous eddy fields
accompanying the anomalous near-surface baroclinicity patterns exhibit,
broadly speaking, structures anticipated from simple linear theories of
baroclinic instability, and suggest a tendency for anomalous wave fluxes to
accelerate--decelerate the surface westerly accordingly. However, the
relationship between anomalous eddy fields and anomalous near-surface
baroclinicity in the midwinter is not consistent with the simple linear
baroclinic instability theories. Composited anomalous SST accompanying 
anomalous near-surface baroclinicity often exhibits
moderate values and large spatial scales in the basin, rather than large
values concentrated near the oceanic fronts. In the midsummer and in some
cases in cold months, however, large SST anomalies are found around the
Kuroshio--Oyashio Extensions. Accompanying anomalies in the net surface
heat flux, SST in the preceding and following months, and meridional eddy
heat flux in the lower troposphere suggest active roles played by the ocean
in generating the concomitant anomalous large-scale atmospheric state in some
of these cases.}

\begin{document}

\maketitle

\statement
Optional significance statement goes here.

\section{Introduction}

It has now become our basic
knowledge that the extratropical
atmosphere is driven strongly by the horizontal
potential temperature gradient that arises from the
differential solar heating (e.g., Lorenz 1955). The horizontal
gradient in the potential temperature, often referred
to as baroclinicity, is a measure of upper-level wind
steering via thermal wind and a measure of baroclinic
instability in the atmosphere. Baroclinicity in the lower
atmosphere in classic theories of atmospheric stability is
measured by a combination of the static stability and
horizontal temperature gradient, the latter of which is
equivalent to vertical shear in the horizontal wind through
the thermal wind balance (Charney 1947; Eady 1949). In
its original form, the Eady's maximum growth rate for
baroclinic instability $B_{GRMax}$ is defined by $B_{GRMax} =
0.31(|f|/N)(\partial U/\partial_z)$ in a zonally homogeneous steady
mean state, where $U$ is the mean zonal flow, $f$ is the
Coriolis parameter, and $N$ is the Brunt--V\"ais\"al\"a frequency.
Charney's formula is slightly different from the Eady's,
but still incorporates the same effects.

Lindzen and Farrell (1980) first applied the Eady's parameter
to atmospheric data to successfully estimate the
maximum growth rate of baroclinic instability in the troposphere.
Hoskins and Valdes (1990) used its localized
version (i.e., $U$, $N$, and $f$ are all local Eulerian mean values)
as the central parameter in their study of the Northern
Hemispheric storm tracks. This local version, or its simplified
version, has been used successfully as an indicator
of baroclinic wave generation in diagnostic studies of
stormtracks in recent years as well (Nakamura and Sampe
2002; Nakamura and Shimpo 2004; Nakamura et al. 2004).
In our study, the North Atlantic part of which was reported
in Nakamura and Yamane (2009, hereafter Part I), we
define the near-surface baroclinic vector, $\mathbf{B}=B^x\mathbf{i} + 
B^y\mathbf{j}$,
where $B^x= -(g/\theta N)(\partial\theta/\partial y)$ 
and $B^y = (g/\theta N)(\partial\theta/\partial x)$  with
$\theta$ being the monthly-mean potential temperature at 2 m
above the surface, and use it as the central quantity of the
diagnoses. Unless stated otherwise, ``anomalies'' refer to
deviations from the climatology hereafter. Though its meridional
component does not appear in any classic theory of
baroclinic instability, a theory that does incorporate the
effect of $B^y$ shows its important role in enhancing baroclinic
wave generation locally to the east of the mean trough
(Niehaus 1980). In the North Atlantic storm track region,
we indeed found that the substantial zonal gradient in the
surface temperature in and around the Labrador Sea plays
a major role in the large-scale atmospheric state.

\section{The sea surface temperature}
The sea surface temperature
 (SST) is an important factor
in determining $\mathbf{B}$ in the storm-track regions (e.g.,
Hoskins and Valdes 1990; Nakamura et al. 2004; Part I).
SST anomalies (SSTAs) around an oceanic front along the
Gulf Stream (GS), Kuroshio Extension (KE), or Oyashio
Extension (OE) can have a profound impact on $\mathbf{B}$ along
the storm tracks.

\subsection{A subtle point}
A subtle but important point that has to
be considered carefully in this regard is the spatial scale
and the location of SSTAs with respect to the climatology,
since it is the anomalous surface temperature
gradient whose structure has a spatial scale of the atmospheric
Rossby deformation radius that can exert significant
influence on the large-scale atmospheric flow.

\subsubsection{Changes in the temperature}
The
high sensitivity of $\mathbf{B}$ to changes in the temperature contrast
across the front and changes in the width of the front, and
the uncertainty in the impact of SSTAs of small spatial
scales on $\mathbf{B}$ make it difficult to assess the effective B
anomalies that are attributable to the SSTAs from the
available data. Moreover, it is uncertain exactly how the
SSTAs in the presence or absence of the land surface
temperature anomalies may or may not produce $\mathbf{B}$ anomalies
that are significant to the atmosphere.

\paragraph{Complicating factor}The complicating
factor introduced by the land surface must be taken
into account when studying potential roles of extratropical
SSTAs in the extratropical atmospheric anomalies.

Lau (1988) investigated patterns of anomalous storm
track activity and associated low-frequency flow anomalies
by computing empirical orthogonal functions (EOFs)
for high-frequency 500-hPa geopotential height for the
Northern Hemisphere winters. He found that both North
Atlantic and North Pacific storm tracks have a pattern of
meridional shift and a pattern of increased or decreased
eddy activity in the first two EOFs.He also found that these
changes in the storm tracks have symbiotic relationships
with the background flows and have substantial impacts on
the mean flow. Part I approached the issue of the stormtrack
and low-frequency flow variability in connection with
SSTAs in the extratropics, focusing on $\mathbf{B}$ as the key parameter
of diagnoses, and found similar patterns of variability
in the eddy activity and low-frequency flow in the
North Atlantic basin. Much of this variability was connected
to SSTAs in the vicinity of the Gulf Stream in cold
months (Part I). Since the winter North Pacific basin has a
storm track andmean flow that appear to be related to the
oceanic fronts, Kuroshio--Oyashio Extensions (KOE) in
this case, in a manner essentially the same as those in the
North Atlantic basin related to the GS,we have attempted
to find similar results for the North Pacific basin. In this
regard, we have chosen not to project our results onto the
major mode of variability in the extratropical North Pacific
basin, the North Pacific decadal variability (PDV), so
that our presentation and discussion are mostly confined
to the wave--mean flow dynamics of monthly time scale or
shorter.

Our approach to the search for a link between anomalies
in the KOE and the overlying atmosphere is as follows:
(i) identify dominant patterns in anomalous $\mathbf{B}$ in the
storm track for each calendarmonth and identify years in
which the anomaly fits the pattern well, (ii) composite
anomalies in the monthly-mean circulation and highfrequency
transients in the atmosphere to obtain a typical
atmospheric state that accompanies the patterns of
anomalous $\mathbf B$, (iii) composite SSTA to obtain a typical
oceanic state that accompanies and precedes the patterns
of anomalous $\mathbf B$, and (iv) composite anomalous net surface
heat flux that accompanies and precedes the pattern of
anomalous $\mathbf B$. With this approach, we obtain typical pictures
of anomalous states in the atmosphere and oceans
with anomalous $\mathbf{B}$ as their connecting interface.
Section 2 describes the data and procedure to compute
$\mathbf B$. Section 3 describes the climatology and variance of $\mathbf B$.
Dominant patterns of $\mathbf B$ are shown in section 4, followed by
composited anomalies in various atmospheric fields and
SST in section 5. Finally, we present our discussion on the
results, examining a potential cause--effect relationship
between anomalies in the SST and atmosphere in section 6.

Figure \ref{fig1} shows the climatology of $Bx$, $U^{200}$,
$\overline{V'\theta'}^{850}$, $\overline{V'\theta'}^{200}$, 
$MR^{z850}$, $U^{1000}$, SST, and $F_h$ for February and
August as examples of the reference state in the winter
and summer. The numeric superscript indicates the pressure
level in hPa. The seasonal mean is visibly more
diffused in structure, particularly for Bx, than those
shown in Fig.~\ref{fig1}. There are no surprises in the overall
picture of Bx. The regions of large land--sea temperature
contrast and the oceanic fronts show very large Bx in cold
months. The position of Bx maximum in the storm track
region is found along theKOEregion throughout the year.
The zonally elongated band of large Bx in the storm track
is generally wider in cold months than in warm months.
In fact, large Bx values that presumably accompany the
Kuroshio and its extension in the cold months vanish in
the summer (Figs. 1a,b). The seasonal variation in the Bx
values in the storm track basically follows that of the
north--south differential heating--largest in the winter and
smallest in the summer. As reported by Nakamura et al.
(2004), the $U^{200}$ maximum is displaced southward from the
Bx maximum visibly in winter months (Fig.~1a), although
U200 is generally large over the area of large Bx in the core
of the storm track. The southward displacement of the
U200 maximum from the Bx maximum in the storm track
is less pronounced or even reversed in warmer months
(Fig.~1b). We note that the structure of the climatological
Bx generally reflects those of ($\partial{\theta^{2{\textrm m}}}/\partial{y}$) with some exceptions where
the surface slope contributes significantly to Bx
in isolated areas over the land, most notably around the
Himalayas.

The position and structure of the storm track as indicated
by $\overline{V'\theta'}^{850}$
and $\overline{V'V'}^{200}$
are, at least for the winter,
essentially the same as those reported in earlier studies on
the storm tracks (e.g., Chang et al. 2002). The maxima in
$\overline{V'\theta'}^{850}$
and $\overline{V'V'}^{200}$ are located in the band of large Bx.

\section{Data and calculation procedures}
The data used to calculate $\mathbf{B}$ are the monthly-mean
temperature at 2 m above the surface ($T^{2m}$) and temperature
at pressure levels available from the 40-yr European
Centre for Medium-Range Weather Forecasts (ECMWF)
Re-Analysis (ERA-40; Uppala et al.~2005). We chose
the ERA-40 $T^{2m}$ data rather than the National Centers
for Environmental Prediction--National Center for Atmospheric
Research (NCEP--NCAR) reanalysis products
for its explicit inclusion of the observed near-surface temperature
in producing the $T^{2m}$ data. The monthly-mean
surface pressure data from the NCEP--NCAR reanalyses
(Kalnay et al. 1996) were used to determine the pressure
levels to be used for $\mathbf B$ calculation, and to calculate $\theta$ 
at 2 m
above the surface from $T^{2m}$. We used the NCEP--NCAR
surface pressure data for convenience, since we had already
compiled the dataset for calculating transient eddy
fluxes to be mentioned later and the ERA-40 surface
pressure data are not readily available. We later compared
the NCEP--NCAR monthly-mean sea level pressure with
that of ERA-40, and found the difference between the two
products to be immaterial for the purpose of the current
study. We also used ERA-40 monthly-mean horizontal
wind and geopotential height at pressure levels, net surface
heat flux, Fh (the sum of latent heat flux, sensible heat flux,
solar radiation, and the thermal radiation), and Hadley
Centre sea surface temperature data (Rayner et al. 2003)
to compile anomaly composites accompanying anomalous
patterns in $\mathbf B$. In addition, we used 6-hourly temperature
and wind data from the NCEP--NCAR reanalyses to
compute various eddy fields. The accuracy of the Fh data
used here is, as true for other reanalyses surface heat flux
products, may not be so high to produce reliable anomaly
composites.

We computed $\mathbf{B}$ near the surface by calculating the
horizontal gradient in $\theta^{2{\textrm m}}$, using the centered finite differencing,
and calculating $N$ from the lowest three vertical
pressure levels that are location dependent because of
topography. Both $\nabla\theta^{2{\textrm m}}$ and $N$ were calculated locally as in
Hoskins and Valdes (1990) and Nakamura and Shimpo
(2004). The entire 45 yr from September 1957 to August
2002 were used for the Northern Hemisphere. To resolve
the dominant modes in $\mathbf{B}$ arising from the land--sea temperature
contrast, one may need much higher horizontal
resolution in the data. The relatively coarse horizontal
resolution of the data may artificially suppress the significance
of the variability associated with the land--sea temperature
contrast.One should keep this limitation in mind.
The 6-hourly bandpassed (period of 2--7 days) eddy
fields and ultra-low-frequency (period of 30 days and longer)
background fields were computed from the NCEP--
NCAR reanalyses, using simple time filters (Lau and Lau
1984) first. The filtered time series were then visually examined
against the raw time series and, then, used to calculate
the slowly evolving bandpassed meridional velocity
variance ($\overline{V'V'}$), meridional temperature flux ($\overline{V'\theta'}$), and
the three-dimensional transient wave activity flux defined
on a zonally varying basic state by Plumb (1986). The wave
activity flux consists of the zonal and meridional advective
fluxes (MU and MV), the zonal and meridional
radiative fluxes (MRx and MRy), and the radiative vertical
flux (MRz). The flux is essentially the Eliassen--Palm
flux (Eliassen and Palm 1961) in a zonally inhomogeneous
mean flow (Plumb 1986). The wave activity flux was calculated
from February 1948 to November 2004 only for
the extratropics poleward of 208 latitude. Also, it was
calculated only from 850 to 30 hPa because of the double
differentiation with respect to pressure required for the
calculation. The flux of particular interest in this study is
the vertical component. Here MR$^z$ is defined by
\[
MR^z=
\frac{
pf \cos\phi}
{p^0|\nabla h\bar q|(d\theta_0/dz)}
\left(\frac{\partial\bar q}{\partial x}
\overline{U'\theta'} +
\frac{\partial\bar q}{\partial y}
\overline{V'\theta'}\right), 
\]
where $q$ is quasigeostrophic potential vorticity, $p$ is the
pressure, $p^0$ is the reference surface pressure set to
1000 hPa here, $\theta_0$ is the area-weighted ultra-low-frequency
hemispheric mean potential temperature at each height,
u9 is the bandpassed potential temperature, and z is the
geopotential height. An overbar denotes an ultra-low frequency
component and a prime denotes bandpassed
component. The 6-hourly time series of wave fluxes was
computed by using the time series of ultra-low-frequency
fields as the basic-state and high-frequency fields as eddies.
In short, the time series was calculated by changing the
meaning of an overbar from the time mean state to an
ultra-low-frequency state, and changing the meaning of a
prime from a departure from the mean to a high-frequency
state. The 6-hourly eddy time series was averaged over each
month to produce monthly-mean time series. This dataset
allows us to examine anomalous eddy fields accompanying
anomalous $\mathbf B$ in specific months. The climatology for
the eddy fields was computed from 46 yr, January 1958 to
December 2003. The calculation of the wave activity and
its flux is described in detail by Nakamura et al. (2010).

\section{Climatology and variance}
The climatology and variance of $Bx$ and $By$ were computed
for each calendar month and examined closely for
their spatial and temporal structures. The monthly climatology,
rather than the seasonal climatology, is used as the
reference in our study, to avoid contamination of the diagnostic
results arising from differences in the climatology
between two successive months. Unlike in the North Atlantic
basin reported in Part I, we found the impact of $By$
variations on the large-scale atmospheric state in the
North Pacific basin much weaker than that of Bx variations,
presumably because of themore zonal orientation of
the KOE in comparison to the Gulf Stream and North
Atlantic Current. In the following, thus, we focus our
presentation on the climatology and variations of $Bx$ and
their impact on the large-scale atmospheric state.

Sample citations: \citet{Becker+Schmitz2003}, \citet{Knutti2008},
and \citep{MeixnerEA2002,Kuji_Nakajima2002,EmeryEA1986}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments
We thank two anonymous reviewers for their comments, which helped
to improve the manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DATA AVAILABILITY STATEMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\datastatement
Start data availability statement here..

%%%%%%%%%%%%%%%%%
%APPENDIXES
%%%%%%%%%%%%%%%%%
\appendix
\appendixtitle{Appendix Title}

\subsection*{Appendix section head}

Here is a sample appendix.
\begin{equation}
\frac{
pf \cos\phi}
{p^0|\nabla h\bar q|(d\theta_0/dz)}
\end{equation}

\appendix[B]
\appendixtitle{Second Appendix Title}
\subsection{Sample appendix section head}
Second appendix example.
\subsection{Sample appendix section head}
\begin{equation}
\left(\frac{\partial\bar q}{\partial x}
\overline{U'\theta'} +
\frac{\partial\bar q}{\partial y}
\overline{V'\theta'}\right) 
\end{equation}

%%%%%%%%%%%%%%%%%
%REFERENCES
%%%%%%%%%%%%%%%%%

\bibliographystyle{ametsoc2014}
\bibliography{references}

%%%%%%%%%%%%%%%%%
% TABLES
%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Percentage of variance explained by the first four
EOFs for the North Pacific Bx. The degree of separation between
EOF1 and EOF2 and EOF2 and EOF3, based on the North et al.
(1982) criterion, is indicated by good (GD) and not good or marginal
(NG).}
\begin{tabular*}{\hsize}{@{\extracolsep\fill}lcccccc@{}}
\topline
Month& EOF1& Split &EOF2& Split& EOF3& EOF4\\
\midline
\ Jan& 29& NG& 24& GD& 10& 5\\
\ Feb& 39& GD& 20& GD& \phantom{1}7 &6\\
\ Mar& 31& GD& 14& NG& 10& 6\\
\ Apr& 23& GD& 14& NG& 10& 7\\
\ May& 19& GD& 12& NG& 10& 7\\
\ Jun& 19& GD& 12& NG& 10& 9\\
\ Jul& 18& NG &13& NG& \phantom{1}9& 7\\
\ Aug& 18& NG& 13& NG& 11& 9\\
\ Sep &17& NG& 13& NG& 10& 8\\
\ Oct &16& NG& 13& GD& \phantom{1}8& 7\\
\ Nov &19 &NG& 16& NG& 11& 8\\
\ Dec& 33& GD& 18& GD& 10& 6\\
\botline
\end{tabular*}
\end{table}

\begin{table}
\centering
\caption{Years selected for anomaly composites for the positive
phase of $B^x$ EOF1.}
\begin{tabular}{lc}
\topline
Month& Yr of positive phase\\
\midline
\ \ Jan& 1961, 1969, 1978, 1979, 1988, 1990, 1992, 1994\\
\ \ Feb& 1964, 1977, 1978, 1980, 1983, 1986, 1988, 2000, 2001\\
\ \ Mar& 1970, 1973, 1979, 1980, 1984, 1988, 2000\\
\ \ Apr& 1959, 1961, 1962, 1963, 1968, 1972, 1983, 2002\\
\ \ May& 1971, 1984, 1993, 1996, 2000\\
\ \ Jun& 1981, 1983, 1984, 1993, 1998\\
\ \ Jul& 1961, 1972, 1973, 1978, 1994, 2000\\
\ \ Aug& 1967, 1970, 1973, 1978, 1994, 1999\\
\ \ Sep& 1975, 1977, 1988, 1989, 1994, 1998, 1999\\
\ \ Oct& 1962, 1977, 1998, 1999, 2001\\
\ \ Nov& 1985, 1986, 1987, 1988, 1991, 1998\\
\ \ Dec& 1957, 1968, 1972, 1978, 1979, 1990\\
\botline
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Years selected for anomaly composites for the negative
phase of $B^x$ EOF1.}
\begin{tabular}{lc}
\topline
Month& Yr of negative phase\\
\midline
\ \ Jan&1962, 1963, 1968, 1974, 1991, 1996, 1997\\
\ \ Feb& 1959, 1963, 1971, 1974, 1976, 1979, 1985, 1989, 1990, 1994\\
\ \ Mar& 1963, 1968, 1972\\
\ \ Apr& 1984, 1988, 1993, 1996, 1999, 2000\\
\ \ May& 1961, 1963, 1983, 1987, 1997\\
\ \ Jun& 1961, 1972, 1978, 1980, 1982, 1986\\
\ \ Jul& 1964, 1974, 1980, 1983, 1986, 1988, 1993\\
\ \ Aug& 1980, 1987, 1991, 1993, 1997\\
\ \ Sep& 1959, 1963, 1969, 1983, 1985, 1993\\
\ \ Oct& 1957, 1986, 1993, 1996, 1997\\
\ \ Nov& 1958, 1968, 1970, 1982, 1997\\
\ \ Dec& 1969, 1974, 1976, 1985, 1998, 1999, 2000, 2001\\
\botline
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%
% FIGURES
%%%%%%%%%%%%%%%%%

\begin{figure}[t]
\centerline{\includegraphics[width=\textwidth]{FigOne.pdf}}

\caption{Climatology of $Bx (10^{-6} s^{-1}$, color) and $U^{200}$(m s$^-1$,
contours) for (a) February and (b) August; $\overline{V'\theta'}^{850}$
(K m s$^-1$, color) and 
$\overline{V'V'}^{200}$
(m$^2$ s$^{-1}$, contours) for (c) February and (d) August; MR$^{z850}$ 
(10$^{-3}$ m$^2$ s$-2$, color) and $U^{1000}$ (m s$^{-1}$, contours) 
for (e) February and (f) August;
and SST (K, color) and $F_h$ [10$^5$ J m$^{-2}$ (6 h)$^{-1}$] for (g)
February and (h) August. Red rectangles indicate the domain of EOF
calculations.} \label{fig1} 
\end{figure}

\begin{figure}[p]
\centerline{\includegraphics{FigTwo.pdf}}
\caption{As in Fig.~10, but for (a),(b) September EOF1; (c),(d) September EOF2; (e),(f) October EOF1; (g),(h) October EOF2; and
(i),(j) December EOF2.}
\end{figure}

\begin{figure}
 \centerline{\includegraphics[width=19pc]{figure01.pdf}}
\appendcaption{A1}{Here is an appendix, single column figure caption.}
\end{figure}

\end{document}

### 2025-04-16 proposal.tex
Initial Overleaf Import

--- 2025-04-16 Examples/amspaperV5.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% amspaperV5.tex --  LaTeX-based template for submissions to American 
-% Meteorological Society journals
-%
+% TemplateV5.tex --  LaTeX-based template for submissions to the 
+% American Meteorological Society
 %
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %%% To be entered by author:
@@
 
-    \authors{Author One\correspondingauthor{Author name, email address}
- and Author Two\thanks{Current affiliation: NCAR, Boulder, Colorado.}}
+    %\authors{Author One\correspondingauthor{Author name, email address}
+% and Author Two\thanks{Current affiliation: American Meteorological Society, 
+    % Boston, Massachusetts.}}
 
-     \affiliation{American Meteorological Society, 
-     Boston, Massachusetts}
+\authors{Author One\correspondingauthor{Author name, email address}}
 
-    \extraauthor{Extra Author}
-    \extraaffil{Affiliation, City, State/Province, Country}
+%% Follow this form:
+    % \affiliation{American Meteorological Society, 
+    % Boston, Massachusetts}
+
+\affiliation{}
+
+%% If appropriate, add additional authors, different affiliations:
+    %\extraauthor{Extra Author}
+    %\extraaffil{Affiliation, City, State/Province, Country}
+
+%\extraauthor{}
+%\extraaffil{}
+
+%% May repeat for a additional authors/affiliations:
+
+%\extraauthor{}
+%\extraaffil{}
 
@@
 %
-% Enter your Abstract here
+% Enter your abstract here
+% Abstracts should not exceed 250 words in length!
+%
 
-\abstract{Enter the text of your abstract here.  This is a sample American
-Meteorological Society (AMS) \LaTeX\ template.  This document provides
-authors with instructions on the use of the AMS \LaTeX\ template.  Authors
-should refer to the file amspaper.tex to review the actual \LaTeX\ code used
-to create this document. The template.tex
-file should be modified by authors for their own manuscript.}
+\abstract{Enter the text of your abstract here.}
 
@@
 %
-\section{Introduction}
 
-This document will provide authors with the basic American Meteorological
-Society (AMS) formatting guidelines. This document was created using \LaTeX\
-and demonstrates how to use the \LaTeX\ template when submitting a manuscript
-to the AMS.  The following sections will outline the guidelines and
-formatting for text, math, figures, and tables while using \LaTeX\/ for a
-submission to the AMS. An attempt to compile amspaper.tex should be made
-before using the template. The files have been tested on Windows, Linux, and
-Mac OS using \TeX\ Live 2019 (available online at
-\url{http://www.tug.org/texlive/}). Feedback and questions should be sent to
-latex@ametsoc.org. Additional information is available on the AMS \LaTeX\ Submission Info
-web page (\url{http://www2.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/author-resources/latex-author-info/}).
+%% In all cases, if there is only one entry of this type within
+%% the higher level heading, use the star form: 
+%%
+% \section{Section title}
+% \subsection*{subsection}
+% text...
+% \section{Section title}
 
-Authors should use the empty template.tex to begin their
-paper. A valuable source of \LaTeX\ information is the {TeX
-Frequently Asked Questions} page (available online at \url{faq.tug.org}).
+%vs
 
-\section{Formatting text and sections}
-The text should be divided into sections, each with a separate heading and
-consecutive numbering. Note, however, that single secondary, tertiary, and
-quaternary sections remain unnumbered. Each section heading should be placed
-on a separate line using the appropriate \LaTeX\ commands.
+% \section{Section title}
+% \subsection{subsection one}
+% text...
+% \subsection{subsection two}
+% \section{Section title}
 
-\subsection*{Secondary headings} 
-Secondary headings labeled with letters are formatted using the
-$\backslash$subsection$*${\tt\string{\string}} for a single
-subsection within a section or
-or $\backslash$subsection{\tt\string{\string}} 
-for multiple
-subsections within one section.
+%%%
+% \section{First primary heading}
 
-\subsubsection*{Tertiary headings} 
-Tertiary headings are formatted using the
-$\backslash$subsubsection$*${\tt\string{\string}} for single a subsubsection
-within a subsection or
-$\backslash$subsubsection{\tt\string{\string}} 
-for multiple subsubsections
-within a subsection.
+% \subsection{First secondary heading}
 
-\paragraph*{Quaternary headings} 
-Quaternary headings are formatted using the
-$\backslash$paragraph$*${\tt\string{\string}} for a single paragraph within
-a subsubsection or
-$\backslash$paragraph{\tt\string{\string}} 
-for multiple paragraphs
-within a subsection.
+% \subsubsection{First tertiary heading}
 
-\section{Citations}
-Citations to standard references in text should consist of the name of the
-author and the year of publication, for example, \citet{Becker+Schmitz2003} or
-\citep{Becker+Schmitz2003} using the appropriate $\backslash$citet\ or
-$\backslash$citep commands, respectively. A variety of citation formats can
-be used with the natbib package; however, the AMS prefers that authors use only the $\backslash$citet\ and
-$\backslash$citep commands. References should be entered in the references.bib file. For a thorough
-discussion of how to enter references into the references.bib database file
-following AMS style, please refer to the \textbf{AMS\_RefsV5.pdf} document
-included in this package.
-
-\section{Formatting math}
-The following sections will outline the basic formatting rules for
-mathematical symbols and units.  In addition, a review of the amspaper.tex
-file will show how this is done with the use of \LaTeX\ commands.  The AMS
-template provides the American Mathematical Society math, font, symbol, and
-boldface packages for use in math mode.
-
-\subsection{Mathematical symbols}
-Symbols must be of the same font style both in text discussion and in
-displayed equations or terms (and figures should be prepared to match).
-Scalar single-character symbols are set italic, Greek, or script. Examples
-are $u$, $L$ [note that $\upsilon$ (Greek upsilon) is used instead of
-\textit{v} (italic ``vee'') to avoid confusion with $\nu$ (Greek nu) often
-used for viscosity; this is handled automatically when in \LaTeX\ math mode],
-$w$, $x$, $y$, $z$, $f$, $g$, $r$, indices such as $i$ or $j$, and constants
-such as $C_D$, $k$, or $K$. Multiple-character scalar variables,
-abbreviations, nondimensional numbers, and acronyms for variables are set
-regular nonitalic: $\mathrm{LWC}$, $\mathrm{Re}$, $\mathrm{Ro}$,
-$\mathrm{BT}$, $\mathrm{abs}$, $\mathrm{obs}$, $\mathrm{max}$,
-$\mathrm{min}$, $\mathrm{Re}$/$\mathrm{Im}$ (real/imaginary), etc. For
-vectors, use boldface nonitalic Times Roman as in $\mathbf{V}$, $\mathbf{v}$,
-or $\mathbf{x}$, and $\mathbf{i}$, $\mathbf{j}$, and $\mathbf{k}$ unit
-vectors. Do not use the \LaTeX\ $\backslash$vec command to denote vectors.
-For matrix notation, use nonitalic boldface Arial (or sans serif) font as in
-$\bm{\mathsf{A}}$, $\bm{\mathsf{B}}$, or $\bm{\mathsf{M}}$.  All mathematical
-operator abbreviations/acronyms are set lowercase regular Roman font, except
-$O$ (on the order of): $\sin$, $\cos$, $\tan$, $\tanh$, $\mathrm{cov}$, $\Pr$
-(for probability; note same as Prandtl number), $\mathrm{const}$ (for
-constant), $\mathrm{c.c.}$ (complex conjugate).
-
-\subsection{Units}
-Units are always set on a single line with a space separating the
-denominator, which is set with a superscript $-1$, $-2$, and so on, rather
-than using a slash for ``per.'' Examples are g kg$^{-1}$, m$^2$ s$^{-1}$, W
-m$^{-2}$, g m$^{-3}$, and m s$^{-1}$ (note that ms$^{-1}$ is the unit for
-``per millisecond'').
-
-\subsection{Equations}
-Brief equations or terms set inline in text must be set as a single-line
-expression because page proofs are not double spaced, for example,
-$\rho^{-1}p/x$ or $(1/{\rho})p/x$ or $(a-b)/(c+d)$; that is, use a
-superscript $-1$ for the denominator. In case of a more complicated term or
-equation, it should be set as an unnumbered display equation, such as
-\begin{displaymath} x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  \end{displaymath}
-
-Otherwise, numbered display equations can be entered using the appropriate
-equation command, such as \begin{equation}
-x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  \end{equation}
-
-Lists of equations are punctuated as written English, and commas, semicolons,
-and periods are placed where appropriate. Conjunctions such as ``and,''
-``while,'' ``when,'' or ``for'' are also typically placed before the final
-element in a mathematical phrase, as befits the intended mathematical
-meaning.
-
-\section{Figures and tables}
-The AMS prefers that all figures and tables are placed {\bf at the end of the document} prior to submission. 
-A list of tables and a list of figures will appear near the end of the PDF
-file, before the actual tables and figures. These lists are necessary for submission.
-
-For appendix figures and tables, special commands are needed to manually 
-change the numbering to ensure that each appendix figure or table is numbered 
-as part of the respective appendix and not as a continuation of the main paper. Use the command
-$\backslash$appendcaption\{\} instead of the usual $\backslash$caption\{\} to adjust the 
-numbering; for example, for Table A1, you would use the command \verb+\appendcaption{A1}+.
-
-Note that the normal $\backslash$ref\{\} command cannot be used to cite 
-appendix figures and tables as the numbering will be incorrect. Callouts for appendix figures 
-and tables in the text will need to be written out as plain text, for example, Fig. A1 and Table A1.
-
-\subsection{Figures}
-The insertion of a sample figure (Fig. \ref{f1})  
-and caption is given below (in the .tex document) and at the end of the document. Standard figure sizes are 19 (one column), 
-27, 33, and 39 (two columns) picas.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% FIGURES---PLACE AT END OF DOCUMENT
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[h]
-% \centerline{\includegraphics[width=19pc]{figure01.pdf}}
-%  \caption{Enter the caption for your figure here.  Repeat as
-%  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
-%\end{figure}
-%
-
-\subsection{Tables}
-Each table must be numbered, provided with a caption, and mentioned
-specifically in the text. 
-See below (in the .tex document) and at the end of the document for the formatting of a sample table (Table \ref{t1}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TABLES---PLACE AT END OF DOCUMENT
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{table}[h]
-%\caption{This is a sample table caption and table layout.  
-%Table from Lorenz (1963).}\label{t1}
-%\begin{center}
-%\begin{tabular}{ccccrrcrc}
-%\topline
-%$N$ & $X$ & $Y$ & $Z$\\
-%\midline
-% 0000 & 0000 & 0010 & 0000 \\
-% 0005 & 0004 & 0012 & 0000 \\
-% 0010 & 0009 & 0020 & 0000 \\
-% 0015 & 0016 & 0036 & 0002 \\
-% 0020 & 0030 & 0066 & 0007 \\
-% 0025 & 0054 & 0115 & 0024 \\
-%\botline
-%\end{tabular}
-%\end{center}
-%\end{table}
-%%
+% \paragraph{First quaternary heading}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Use \appendix if there is only one appendix.
+%\appendix
 
-%% If only one appendix, use
-%%\appendix%
+% Use \appendix[A], \appendix[B], if you have multiple appendixes.
+%\appendix[A]
 
-%% If more than one appendix, use \appendix[<letter>], e.g.,
- \appendix[A]
+%% Appendix title is necessary! For appendix title:
+%\appendixtitle{}
 
-\appendixtitle{Title of Appendix}
+%%% Appendix section numbering (note, skip \section and begin with \subsection)
+% \subsection{First primary heading}
 
-\subsection*{Appendix section}
+% \subsubsection{First secondary heading}
 
-The AMS template allows authors to format an unlimited number of
-appendixes. To format a single appendix, use the $\backslash$appendix
-command with no additional argument. Otherwise, add the appropriate
-one-letter argument to the $\backslash$appendix command (e.g.
-$\backslash$appendix[A], $\backslash$appendix[B],
-$\backslash$appendix[C], etc.) corresponding to the appropriate
-appendix.
+% \paragraph{First tertiary heading}
 
-The title of the appendix can be formatted using the
-$\backslash$appendixtitle{\tt\string{\string}} 
- command. The $\backslash$subsection, $\backslash$subsubection,
-and $\backslash$paragraph commands are used to create sections within
-the appendix. (Note that the appendix title takes the place of $\backslash$section 
-in the appendix, so the first section should begin with $\backslash$subsection
-instead of $\backslash$section.)
- Equations are automatically numbered appropriately for 
-each appendix. Here is an example of the first equation in appendix
-A, automatically labeled (\ref{eq:1}): 
-\begin{equation} \label{eq:1}
-x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
-\end{equation}
-
-For appendix figures and tables, special commands are needed to manually 
-change the numbering to ensure that each appendix figure or table is numbered 
-as part of the appendix and not as a continuation of the main paper. Use the command
-$\backslash$appendcaption\{\} instead of the usual $\backslash$caption\{\} to adjust the 
-numbering; for example, for Table A1, you would use the command \verb+\appendcaption{A1}+.
-In-text callouts for each appendix figure and table will need to be written as plain text;
-the usual $\backslash$ref\{\} command cannot be used.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%APPENDIX FIGURE AND TABLE EXAMPLES---PLACE AT END OF DOCUMENT
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%\begin{table}
-%\appendcaption{A1}{Here is the appendix table caption.}
-%\centering
-%\begin{tabular}{ccc}
-%a&b&c\\
-%d&e&f
-%\end{tabular}
-%\end{table}
+%% Important!
+%\appendcaption{<appendix letter and number>}{<caption>} 
+%must be used for figures and tables in appendixes, e.g.,
 %
 %\begin{figure}
-%\centerline{(illustration here)}
-%\appendcaption{A1}{Here is the appendix figure caption.}
+%\noindent\includegraphics[width=19pc,angle=0]{figure01.pdf}\\
+%\appendcaption{A1}{Caption here.}
 %\end{figure}
 %
-
-\appendix[B]
-\appendixtitle{File Structure of the AMS \LaTeX\ Package}
-
-\subsection{AMS \LaTeX\ files}
-You will be provided with a tarred, zipped \LaTeX\ package containing 
-17 files. These files are
-
-\begin{description}
-\item[Basic style file:] ametsocV5.cls.
-
-The file ametsocV5.cls is the manuscript style file.
-
-\begin{itemize}
-\item
-Using \verb+\documentclass{ametsocV5}+ for your .tex document
-will 
-generate a PDF that follows all AMS guidelines for submission and peer
-review.
-
-\item
-Using \verb+\documentclass[twocol]{ametsocV5}+ for your .tex document
-can be used to generate a PDF that closely
-follows the layout of an AMS journal page, including single spacing and two
-columns.  This journal style PDF is only for the author's personal use, and
-any papers submitted in this style will not be accepted.  
-\end{itemize}
-Always use \verb+\documentclass{ametsocV5}+ 
-when generating a PDF for submission to the AMS.
-
-\item[Template:]
-templateV5.tex, for the author to use when making his/her
-paper.
-The file provides a basic blank template with some
-section headings for authors to easily enter their manuscript.
-
-\item[Sample .tex and .pdf files:]
-The file amspapeV5r.tex contains the \LaTeX\ code for the sample file.  
-The resulting PDF can be seen in amspaper.pdf (this file).
-
-\item[Sample article:] article formatted in draft and two-column mode.
-
-\begin{itemize}
-\item
-AMSSamp1V5.tex, AMSSamp1V5.pdf\\
-Formal paper done in draft mode and the resulting .pdf.
-
-\item
-AMSSamp2V5.tex, AMSSamp2V5.pdf \\
-The same paper using the \verb+[twocol]+ option and the resulting .pdf.
-
-\item
-FigOne.pdf, FigTwo.pdf, and figure01.pdf are sample figures.
-
-\end{itemize}
-
-\item[Bibliography Files:]
-
-ametsoc2014.bst, database2020.bib, and references.bib.
-
-\begin{itemize}
-\item
-ametsoc2014.bst is the bibliography style file.
-
-\item
-database2020.bib is an example of a bibliographic database file.
-
-\item
-references.bib should be altered with your own bibliography information.  
-\end{itemize}
-
-\item[Documention:] found in AMSDocsV5.pdf. Additional information
-found in
-readme.txt, which contains a list of the files and how they are used.
-
-\end{description}
-
-\subsection{Help for Authors}
-Questions and feedback concerning the use of the AMS \LaTeX\ files should be
-directed to latex@ametsoc.org. Additional information is available on the AMS
-\LaTeX\ Submission Info web page (\url{http://www2.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/author-resources/latex-author-info/}).
-
-\appendix[C]
-\appendixtitle{Building a PDF and Submitting Your 
-\LaTeX\ Manuscript Files to the AMS}
-
-\subsection{Building your own PDF}
-There are a variety of different methods and programs that will create a
-final PDF from your \LaTeX\ files. The easiest method is to download one of
-the freely available text editors/compilers such as TexWorks or TeXnicCenter.
-TexWorks is installed with the TeXLive distribution and provides both a text
-editor and the ability to compile your files into a PDF.
-
-\subsection{Submitting your files to the AMS for peer review}
-The AMS uses the Editorial Manager system for all author submissions for
-peer review. Editorial Manager uses the freely available \TeX\ Live 2018
-distribution. This system will automatically generate a PDF from your
-submitted \LaTeX\ files and figures.
-
-You should not upload your own PDF into
-the system. If the system does not build the PDF from your files correctly,
-refer to the AMS \LaTeX\ FAQ page first for possible solutions. If your PDF
-still does not build correctly after trying the solutions on the FAQ page, email
-latex@ametsoc.org for help.
-
-\subsection{Other software}
-As mentioned above, there is a variety of software that can be used to edit
-.tex files and build a PDF.  The AMS does not support \LaTeX\/-related
-WYSIWYG software, such as Scientific Workplace, or WYSIWYM software, such as
-LyX.  \TeX\ Live (available online at \\ \url{http://www.tug.org/texlive/}) is
-recommended for users needing an up-to-date \LaTeX\ distribution with
-software that includes an editor and the ability to automatically generate a
-PDF.
-
+% All appendix figures/tables should be placed in order AFTER the main figures/tables, i.e., tables, appendix tables, figures, appendix figures.
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
- This shows how to enter the commands for making a bibliography using
- BibTeX. It uses references.bib and the ametsoc2014.bst file for the style.
-
-\bibliographystyle{ametsoc2014}
-\bibliography{references}
+% Make your BibTeX bibliography by using these commands:
+% \bibliographystyle{ametsoc2014}
+% \bibliography{references}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[h]
-\caption{This is a sample table caption and table layout.}\label{t1}
-\begin{center}
-\begin{tabular}{ccccrrcrc}
-\topline
-$N$ & $X$ & $Y$ & $Z$\\
-\midline
- 0000 & 0000 & 0010 & 0000 \\
- 0005 & 0004 & 0012 & 0000 \\
- 0010 & 0009 & 0020 & 0000 \\
- 0015 & 0016 & 0036 & 0002 \\
- 0020 & 0030 & 0066 & 0007 \\
- 0025 & 0054 & 0115 & 0024 \\
-\botline
-\end{tabular}
-\end{center}
-\end{table}
+%% Enter tables at the end of the document, before figures.
+%%
 %
-
-\begin{table}
-\appendcaption{A1}{Here is the appendix table caption.}
-\centering
-\begin{tabular}{ccc}
-\topline
-$1$ & $2$ & $3$ \\
-\midline
-a&b&c \\
-d&e&f \\
-\botline
-\end{tabular}
-\end{table}
+%\begin{table}[t]
+%\caption{This is a sample table caption and table layout.  Enter as many tables as
+%  necessary at the end of your manuscript. Table from Lorenz (1963).}\label{t1}
+%\begin{center}
+%\begin{tabular}{ccccrrcrc}
+%\hline\hline
+%$N$ & $X$ & $Y$ & $Z$\\
+%\hline
+% 0000 & 0000 & 0010 & 0000 \\
+% 0005 & 0004 & 0012 & 0000 \\
+% 0010 & 0009 & 0020 & 0000 \\
+% 0015 & 0016 & 0036 & 0002 \\
+% 0020 & 0030 & 0066 & 0007 \\
+% 0025 & 0054 & 0115 & 0024 \\
+%\hline
+%\end{tabular}
+%\end{center}
+%\end{table}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
- \centerline{\includegraphics[width=19pc]{figure01.pdf}}
-  \caption{Enter the caption for your figure here.  Repeat as
-  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
-\end{figure}
-
-\begin{figure}
-\centerline{(illustration here)}
-\appendcaption{A1}{Here is the appendix figure caption.}
-\end{figure}
-
-\begin{figure}
-\centerline{(illustration here)}
-\appendcaption{B1}{Here is the appendix figure caption.}
-\end{figure}
+%% Enter figures at the end of the document, after tables.
+%%
+%
+%\begin{figure}[t]
+%  \noindent\includegraphics[width=19pc,angle=0]{figure01.pdf}\\
+%  \caption{Enter the caption for your figure here.  Repeat as
+%  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
+%\end{figure}
 
 \end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% END OF AMSPAPER.TEX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-04-16 forecast_robustness.R
Added initial R file

@@ -1,0 +1,119 @@
+
+library(dplyr)
+library(RJSONIO)
+library(curl)
+library(abind)
+library(sp)
+library(tidyr)
+library(httr)
+library(ggplot2)
+
+
+month_names <- c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec")
+
+### SETUP - CHANGE STUFF HERE ###
+
+## set what data you'd like to download
+
+# country / project
+
+country <- c("niger")
+
+# freqs (0-100)
+
+freqs <- seq(5,95,by=5)
+
+# months (numbered 0-11)
+
+months <- c(0:5) 
+
+months_df <- data.frame("month" = c(0:11),"name" = month_names)
+
+prev_months_df <- data.frame("month" = c(1:11,0),"name" = month_names)
+
+# obs dataset (see config for list of possible options)
+
+obs_dataset <- c("chirps-precip-jas")
+
+# server (e.g. production or test)
+
+server <- "https://iridl.ldeo.columbia.edu/fbfmaproom2/"
+
+## get key guide
+
+# admin 0
+
+admin_0_key <- bind_rows(fromJSON(paste0(server,"regions?country=",country,"&level=0")))
+
+## function to set query paths 
+
+fun_download_path <- function(country,level,region,month,freq,predictor,predictand){
+  
+  query_path <- paste0(server,country,"/export?",
+                       "&mode=",level,
+                       "&region=",region,
+                       "&season=season1&issue_month0=",month,
+                       "&freq=",freq,
+                       "&predictor=",predictor,
+                       "&predictand=",predictand) 
+  return(query_path)
+  
+}
+
+### DOWNLOADS ###
+
+## admin 0
+
+admin_0_downloads <- list()
+
+for (freq in freqs) {
+  
+  admin_0_downloads[[as.character(freq)]] <- list()
+
+  for (month in months) {
+    
+    api_download <- fromJSON(fun_download_path(country,0,admin_0_key[1,1],month,freq,"pnep",obs_dataset[1]),nullValue = NA)
+    admin_0_downloads[[as.character(freq)]][[as.character(month)]] <- api_download
+    
+
+  }
+  
+}
+
+admin_0_bad_years <- fromJSON(fun_download_path(country,0,admin_0_key[1,1],0,freq,"pnep",'bad-years-v2'),nullValue = NA)
+
+
[…]

### 2025-04-17 proposal.tex
abstract draft ks

@@
 
-\abstract{Enter the text of your abstract here.}
+\abstract{As of 2022, nearly 5 million smallholder farmers across a dozen countries are covered by Anticipatory Action (AA) programs \citep{chaves-gonzalez_anticipatory_2022}, which are aid schemes triggered if seasonal forecasts of rainfall cross a locally-determined threshold.
+By compensating farmers when yields are expected to be low, AA programs can help smooth risks, enhance food security, and reduce the likelihood of disaster-induced displacement \citep{thalheimer_forecast-based_2020,XX}.
+AA thresholds are often calculated to meet a donor or government [XX] trigger frequency; for example, [XX]’s AA program intends to trigger in the worst 20\% [XX] of seasons by expected agricultural yield.
+However, the observational rainfall record in many regions with AA programs is limited, and interannual or even interdecadal autocorrelation in the climate patterns that affect seasonal rainfall further reduces the effective sample size used to calculate these thresholds.
+This suggests that AA triggers may be highly sensitive to the available sample size of the observational record, increasing the risk that, due to poorly specified thresholds, either action is not triggered when it is needed or programs become financially unsustainable.
+In practice, forecasters are aware of this uncertainty, and ad hoc decisions on whether to trigger action are often made if seasonal forecasts are close to the threshold.
+To better support decision-making and formalize a process of determining whether a threshold for action may be crossed, we develop locally-defined confidence intervals for AA thresholds by recalculating triggers using bootstrap resampling of the observational record.
+We find that, though [XX], in regions with [XX characteristics/teleconnections/etc.], AA triggers are particularly unstable, [XX raising questions about the suitability of simple thresholds there] [XX alternatively; we find confidence intervals are narrow in most places, supporting the use of simple triggers...]
 
+[XX something something, these triggers are likely to be nonstationary in a changing climate as well. I think \citep{coughlan_de_perez_adapting_2022} says something about that]}
 \begin{document}

### 2025-08-21 proposal.tex

--- 2025-04-16 proposal.tex
@@
 \usepackage{newtxmath}
+\usepackage{natbib}
 
@@
 
-\title{Title here}
+\title{Quantifying Uncertainty in Forecast-Based Anticipatory Action Triggers Using Bootstrapping Methods}
 
@@
 
-\authors{Author One\correspondingauthor{Author name, email address}}
-
-%% Follow this form:
-    % \affiliation{American Meteorological Society, 
-    % Boston, Massachusetts}
-
-\affiliation{}
-
-%% If appropriate, add additional authors, different affiliations:
-    %\extraauthor{Extra Author}
-    %\extraaffil{Affiliation, City, State/Province, Country}
-
-%\extraauthor{}
-%\extraaffil{}
-
-%% May repeat for a additional authors/affiliations:
-
-%\extraauthor{}
-%\extraaffil{}
+\author{%
+Max Mauerman\affmark[1]\correspondingauthor{Max Mauerman, max.mauerman@columbia.edu}, Kevin Schwarzwald\affmark[1], Daniel Osgood\affmark[1], Nathan Lenssen\affmark[2], and Abdel-Lathif Younous\affmark[3]\\
+\affaddr{\affmark[1] Columbia University Climate School}\\
+\affaddr{\affmark[2] Colorado School of Mines \& NSF National Center for Atmospheric Reserach}\\
+\affaddr{\affmark[3] United Nations World Food Programme}\\
+}
 
@@
 
-\abstract{Enter the text of your abstract here.}
-
+\abstract{As of 2022, nearly 5 million smallholder farmers across a dozen countries are covered by Anticipatory Action (AA) programs \citep{chaves-gonzalez_anticipatory_2022}, which are aid schemes triggered if seasonal forecasts of rainfall cross a locally-determined threshold.
+AA thresholds are often calculated to meet a donor- or government-specified expected trigger frequency \citep{coughlan_de_perez_adapting_2022,world_food_programme_10_2024}.
+However, the observational rainfall record in many regions with AA programs is limited, and interannual and interdecadal autocorrelation in the climate patterns driving seasonal rainfall further reduce the effective sample size used to calculate these thresholds \citep{martinez_seasonal_2022}.
+Thus, AA triggers are highly sensitive to the limited observational record, increasing the risk that action is not triggered when it is needed or that programs become financially unsustainable due to excessive triggers.
+In practice, forecasters are aware of this uncertainty, and ad hoc decisions on whether to trigger action are often made if seasonal forecasts are close to the threshold.
+Seeking a more formalized, statistically sound alternative to ad hoc approaches, governments and humanitarian agencies have asked for transparent and intuitive metrics for forecast trigger uncertainty.
+To address this demand, we develop a method for quantifying the uncertainty in thresholds due to a limited observational record using bootstrap resampling.
+Examining United Nations operational AA forecasts in sub-Saharan Africa, we demonstrate how our method provides a better quantification of critical underlying uncertainty, enabling these AA systems to meet the state-of-the-art demanded by donors and the WMO.}
 \begin{document}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MAIN BODY OF PAPER
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-
-%% In all cases, if there is only one entry of this type within
-%% the higher level heading, use the star form: 
-%%
-% \section{Section title}
-% \subsection*{subsection}
-% text...
-% \section{Section title}
-
-%vs
-
-% \section{Section title}
-% \subsection{subsection one}
-% text...
-% \subsection{subsection two}
-% \section{Section title}
-
-%%%
-% \section{First primary heading}
-
-% \subsection{First secondary heading}
-
-% \subsubsection{First tertiary heading}
-
-% \paragraph{First quaternary heading}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% ACKNOWLEDGMENTS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments
-Keep acknowledgments (note correct spelling: no ``e'' between the ``g'' and
-``m'') as brief as possible. In general, acknowledge only direct help in
-writing or research. Financial support (e.g., grant numbers) for the work
-done, for an author, or for the laboratory where the work was performed is
-best acknowledged here rather than as footnotes to the title or to an
-author's name. Contribution numbers (if the work has been published by the
-author's institution or organization) should be included as footnotes on the title page,
-not in the acknowledgments.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DATA AVAILABILITY STATEMENT
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 
-%
-\datastatement
-The data availability statement is where authors should describe how the data underlying 
-the findings within the article can be accessed and reused. Authors should attempt to 
-provide unrestricted access to all data and materials underlying reported findings. 
-If data access is restricted, authors must mention this in the statement.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIXES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Use \appendix if there is only one appendix.
-%\appendix
-
-% Use \appendix[A], \appendix[B], if you have multiple appendixes.
-%\appendix[A]
-
-%% Appendix title is necessary! For appendix title:
-%\appendixtitle{}
-
-%%% Appendix section numbering (note, skip \section and begin with \subsection)
-% \subsection{First primary heading}
-
-% \subsubsection{First secondary heading}
-
-% \paragraph{First tertiary heading}
-
-%% Important!
-%\appendcaption{<appendix letter and number>}{<caption>} 
-%must be used for figures and tables in appendixes, e.g.,
-%
-%\begin{figure}
-%\noindent\includegraphics[width=19pc,angle=0]{figure01.pdf}\\
-%\appendcaption{A1}{Caption here.}
-%\end{figure}
-%
-% All appendix figures/tables should be placed in order AFTER the main figures/tables, i.e., tables, appendix tables, figures, appendix figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % REFERENCES
@@
 % Make your BibTeX bibliography by using these commands:
-% \bibliographystyle{ametsoc2014}
-% \bibliography{references}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TABLES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Enter tables at the end of the document, before figures.
-%%
-%
-%\begin{table}[t]
-%\caption{This is a sample table caption and table layout.  Enter as many tables as
-%  necessary at the end of your manuscript. Table from Lorenz (1963).}\label{t1}
-%\begin{center}
-%\begin{tabular}{ccccrrcrc}
-%\hline\hline
-%$N$ & $X$ & $Y$ & $Z$\\
-%\hline
-% 0000 & 0000 & 0010 & 0000 \\
-% 0005 & 0004 & 0012 & 0000 \\
-% 0010 & 0009 & 0020 & 0000 \\
-% 0015 & 0016 & 0036 & 0002 \\
-% 0020 & 0030 & 0066 & 0007 \\
-% 0025 & 0054 & 0115 & 0024 \\
-%\hline
-%\end{tabular}
-%\end{center}
-%\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% FIGURES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Enter figures at the end of the document, after tables.
-%%
-%
-%\begin{figure}[t]
-%  \noindent\includegraphics[width=19pc,angle=0]{figure01.pdf}\\
-%  \caption{Enter the caption for your figure here.  Repeat as
-%  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
-%\end{figure}
+\bibliographystyle{ametsoc2014}
+\bibliography{references}
 

### 2025-08-21 paper.tex

%% Version 5.0, 2 January 2020
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TemplateV5.tex --  LaTeX-based template for submissions to the 
% American Meteorological Society
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Start with one of the following:
% DOUBLE-SPACED VERSION FOR SUBMISSION TO THE AMS
\documentclass{ametsocV5}

% TWO-COLUMN JOURNAL PAGE LAYOUT---FOR AUTHOR USE ONLY
% \documentclass[twocol]{ametsocV5}

% Enter packages here. If too many math alphabets are used,
% remove unnecessary packages or define hmmax and bmmax as necessary.

%\newcommand{\hmmax}{0}
%\newcommand{\bmmax}{0}
\usepackage{amsmath,amsfonts,amssymb,bm}
\usepackage{mathptmx}%{times}
\usepackage{newtxtext}
\usepackage{newtxmath}
\usepackage{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% To be entered by author:

%% May use \\ to break lines in title:

\title{Quantifying Uncertainty in Forecast-Based Anticipatory Action Triggers Using Bootstrapping Methods}

%%% Enter authors' names, as you see in this example:
%%% Use \correspondingauthor{} and \thanks{Current Affiliation:...}
%%% immediately following the appropriate author.
%%%
%%% Note that the \correspondingauthor{} command is NECESSARY.
%%% The \thanks{} commands are OPTIONAL.

    %\authors{Author One\correspondingauthor{Author name, email address}
% and Author Two\thanks{Current affiliation: American Meteorological Society, 
    % Boston, Massachusetts.}}

\author{%
Max Mauerman\affmark[1]\correspondingauthor{Max Mauerman, max.mauerman@columbia.edu}, Kevin Schwarzwald\affmark[1], Daniel Osgood\affmark[1], Nathan Lenssen\affmark[2], and Abdel-Lathif Younous\affmark[3]\\
\affaddr{\affmark[1] Columbia University Climate School}\\
\affaddr{\affmark[2] Colorado School of Mines \& NSF National Center for Atmospheric Research}\\
\affaddr{\affmark[3] United Nations World Food Programme}\\
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%
% Enter your abstract here
% Abstracts should not exceed 250 words in length!
%

\abstract{As of 2022, nearly 5 million smallholder farmers across a dozen countries are covered by Anticipatory Action (AA) programs \citep{chaves-gonzalez_anticipatory_2022}, which are aid schemes triggered if seasonal forecasts of rainfall cross a locally-determined threshold. AA thresholds are often calculated to meet a donor- or government-specified expected trigger frequency \citep{coughlan_de_perez_adapting_2022,world_food_programme_10_2024}. However, the observational rainfall record in many regions with AA programs is limited, and interannual and interdecadal autocorrelation in the climate patterns driving seasonal rainfall further reduce the effective sample size used to calculate these thresholds \citep{martinez_seasonal_2022}. Thus, AA triggers are highly sensitive to the limited observational record, increasing the risk that action is not triggered when it is needed or that programs become financially unsustainable due to excessive triggers. In practice, forecasters are aware of this uncertainty, and ad hoc decisions on whether to trigger action are often made if seasonal forecasts are close to the threshold. Seeking a more formalized, statistically sound alternative to ad hoc approaches, governments and humanitarian agencies have asked for transparent and intuitive metrics for forecast trigger uncertainty. To address this demand, we develop a method for quantifying the uncertainty in thresholds due to a limited observational record using bootstrap resampling. Examining United Nations operational AA forecasts in sub-Saharan Africa, we demonstrate how our method provides a better quantification of critical underlying uncertainty, enabling these AA systems to meet the state-of-the-art demanded by donors and the WMO.}
\begin{document}

%% Necessary!
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SIGNIFICANCE STATEMENT/CAPSULE SUMMARY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% If you are including an optional significance statement for a journal article or a required capsule summary for BAMS 
% (see www.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/formatting-and-manuscript-components for details), 
% please apply the necessary command as shown below:
%
% \statement
% Significance statement here.
%
% \capsule
% Capsule summary here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAIN BODY OF PAPER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

%% In all cases, if there is only one entry of this type within
%% the higher level heading, use the star form: 
%%
% \section{Section title}
% \subsection*{subsection}
% text...
% \section{Section title}

%vs

% \section{Section title}
% \subsection{subsection one}
% text...
% \subsection{subsection two}
% \section{Section title}

%%%
% \section{First primary heading}

% \subsection{First secondary heading}

% \subsubsection{First tertiary heading}

% \paragraph{First quaternary heading}

\section{Introduction}

% Max to draft and Dan to refine.

Humanitarian agencies are trying to be less reactive... advances in the state of the science on seasonal forecasting make early action against drought feasible... But forecasts need to follow standards of consistency, robustness and transparency a la WMO guidelines, especially when lives are on the line... it is not always obvious how to pragmatically account for uncertainty under those guidelines... our paper advances a simple method for quantifying one type of uncertainty, which is how sensitive the trigger definition is to the limited historical record... this method was developed to respond to practical needs from WFP, and can be replicated elsewhere.

\section{Forecast-based anticipatory action}

\subsection{Institutional background of AA}

% Abdel to help write this part.

When did the UN start using forecasts for humanitarian action? What about seasonal forecasts in particular? What obstacles (scientific and practical) have kept this from happening sooner? Some parallels to similar problems in the area of parametric insurance.

\subsection{The PyCPT seasonal forecasting method}

% Max to write but could probably use Kevin and Nathan's eyes to make sure I'm not saying anything scientifically out of pocket.
% Dan style note: "PyCPT" instead of "NextGen" 
% Talk about heuristic rules that the country offices were using, and how they are similar conceptually / functionallty to bootstrapping

Talk about the types of seasonal forecasts we are using here, where they come from and how to interpret them. Can draw from a lot of prior IRI work here. Key talking points are that NextGen is a method to produce downscaled and bias-corrected seasonal forecasts using a multi-model ensemble of GCMs. The models, predictand and calibration parameters can be customized to each locality and context. Local met services use the method and software to create purpose-built forecast models for their country or region. Forecasts are presented as the probability of non-exceedance at a given percentile threshold (not only terciles).

\subsection{Triggering humanitarian action using seasonal forecasts}

% Max to write with Abdel's input.

Talk about the mechanics of how we define a forecast action trigger here. Below is how I wrote it up for WFP in Ethiopia:

A core component of the Anticipatory Action Plan is the forecast “trigger”, i.e., the quantitative threshold of predictive certainty above which a certain anticipatory action will be taken. 
This document describes the procedure used to determine triggers for the WFP AA project and the rationale for why it was implemented this way. At the end of this document, there are links to a set of live documents which provide the most up to date trigger information for each active region and woreda in the AA project.

The basic rationale behind the choice of triggers is that they should be transparent, consistently applied and easily related to the expected frequency with which action will be taken. A humanitarian agency planning budgets for AA programs should be able to look at the trigger rule associated with a specific action and know that it is likely to occur - for example - 20 percent  of the time, based on historical analysis.

Accordingly, the trigger rule used for this AAP is based on a consistent frequency of action for a given type of programming, across locations and times of year. For instance, in Somali Region, “Severe” level actions are associated with a 20 percent frequency of action. This means that for a given location, for the forecast issued in a given month, we should expect Severe actions to trigger during the strongest 20pct of drought forecasts. To determine the numeric threshold associated with acting 20pct of the time, we rank all of the historical forecasts (“hindcasts”) available from the least likely prediction of drought to the most likely, and compute the 20th percentile of this distribution.

Talk a little about the sensitivity of this method to the choice of threshold, and what to do when the forecast is just a couple pp above or below the threshold. This is something that practitioners always have to deal with, and they have heuristic ways of managing it, but those can be inconsistent. We propose here a more systematic solution... segues into discussion about sources of uncertainty.

\section{Sources of forecast uncertainty}

% Nathan - can fold most of this into section 2b, since it isn't really a standalone section

% Kevin to write w Max and Nathan as needed.
% The basic strategy is to go through all of the different commonly discussed sources of uncertainty, and talk about how they are addressed in the NextGen approach, except for the problem of the limited historical record. That is what we're addressing with this paper.

% note Aug 13: Can pare down the discussion of the first four sources of uncertainty, since they aren't the main point.

\subsection{Uncertainty in modeling choices}

Different GCMs have different assumptions about how things work. We account for this by using a multi-model ensemble in predictions.

\subsection{Uncertainty in initial conditions}

I'm not sure if or how to talk about this -- I'm mostly familiar with this in the context of long-term climate change projections, not seasonal forecasting. As I understand it, all of the GCM output that we use for seasonal forecasting is initialized with the same conditions for a given issue month. So maybe this is not a huge issue the way it is for CC projections, where uncertainty in initial conditions can propagate further (cf Schwarzwald and Lenssen 2022). But maybe I'm also missing something, so it would be good to get your input.

In any case, we can talk about how we use forecasts at different lead times, with the decision rule putting greater weight on forecasts with a shorter lead time, since we are more certain about the initial conditions.

\subsection{Uncertainty in accuracy of predictions}

The bias correction and calibration components of NextGen are meant to account for this. I need to bug Andy R. and Azhar about this since they have promised a more thorough writeup. Talk about the sharpness ratio of quantifying the confidence of the forecast (this is something that decision-makers should also be thinking about, and not only whether the forecast is above or below the threshold).

Note that this section is mostly focusing on how well forecasts predict observed rainfall. There is a second-order question about whether forecasted rainfall is related to humanitarian impacts of drought (e.g. hunger, loss of animals), which is of practical importance to the projects, but not really what this paper is about. We may want to at least mention it in passing.

\subsection{Uncertainty due to the limited historical record}

What we wrote for the abstract: The observational rainfall record in many regions with AA programs is limited, and interannual and interdecadal autocorrelation in the climate patterns driving seasonal rainfall further reduce the effective sample size used to calculate these thresholds \citep{martinez_seasonal_2022}. Thus, AA triggers are highly sensitive to the limited observational record, increasing the risk that action is not triggered when it is needed or that programs become financially unsustainable due to excessive triggers. In practice, forecasters are aware of this uncertainty, and ad hoc decisions on whether to trigger action are often made if seasonal forecasts are close to the threshold. Seeking a more formalized, statistically sound alternative to ad hoc approaches, governments and humanitarian agencies have asked for transparent and intuitive metrics for forecast trigger uncertainty.

\section{The bootstrapping method for limited data}

% Nathan to write w Kevin and Max and needed

% MAX TODO: Make sure we are using block randomization, with consecutive years. This could help deal with the autocorrelation problem. There is a literature we can tie this to

% Abdel is particularly interested in the question of whether interannual correlation and / or trends make the calculation of confidence intervals using this method problematic. i need to put on my stats hat this this one.

% Note that if we were really doing this properly, I think we would want to do the bootstrap sampling during the step of per-pixel model calibration (which happens in CPT), rather than doing it after the fact with areal average data as we are doing here. But that is a matter for a future paper. We want to at least mention it though I think.

% kevin note: we are thinking about this as applicable to things beyond PyCPT, tho, so dont get too hung up about this

Below is what I've written about the method for WFP. We need to motivate why we're using the method of bootstrapping and why it's reasonable in this context first.

The AA protocol is based on a binary trigger - if the drought forecast is above a given threshold of confidence, it triggers; if it is below this threshold, it does not.

To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis. This analysis is based on the method of re-estimating the trigger threshold with years randomly omitted from the historical record. Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error.

In this case, we run 100 iterations of the re-sampling process. In each iteration, 5 years are randomly omitted from the historical record and then the threshold is re-computed.

The box plots below depict the resulting sensitivity range in the trigger threshold for each issue month and severity level. The actual trigger threshold is shown in red. For reference, the climatological odds (the probability of non-exceedence if we were to guess from historical average conditions, without any forecast information) is shown as the dotted line.

If the current forecast is outside of the sensitivity range, this indicates that we have statistically robust evidence that it is above or below the threshold. If the current forecast falls within the sensitivity range, this indicates that we should interpret the trigger result for the current season with caution.

\section{Application to WFP assistance programs}

\subsection{Results}

% Each of these is an active WFP country, and the forecast is for the primary agricultural season in that country, which varies. Need to mention all that stuff along with the model parmeters in an appendix...

% Need to be sure we are using a definition of the percentile that is consistent with what the API uses! R's may not be the same. make sure we are consistent

% add table of country results, but dont get too into the weeds on country specifics

% fig 1: simple obs (NOT forecast) illustration of what we are doing
% fig 2: example using actual WFP data. can just be one area-month-freq. we can put the rest in an appendix.

% tweak visual so that user focuses on the boxplot, not the clim odds

% Note: The figures seem to get added to the end instead of in-line. I'm guessing that's how it is in the journal?

% note that this method can acommodate compound / multimonth rules

\subsubsection{Ethiopia}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/ethiopia.png}
\end{figure}

\subsubsection{Madagascar}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/madagascar.png}
\end{figure}

\subsubsection{Guatemala}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/guatemala-jja.png}
\end{figure}

\subsubsection{Niger}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/niger.png}
\end{figure}

\subsubsection{Djibouti}

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/djibouti-ond.png}
\end{figure}

\subsection{Discussion}

- in general, the range of uncertainty tends to be a couple of percentage points in either direction.
- forecast sharpness generally increases as the lead time decreases, but the threshold uncertainty doesn't.
- likewise, madagascar and guatemala have the sharpest forecasts (which makes some sense; we have reason to think that ENSO teleconnections are relatively stronger there), but have as much or more threshold uncertainy than the other countries.
- this highlights the fact that we have to think about the innate uncertainty from having a short historical record, even in places where we think the forecast model has historically performed well!
- also in some places (ex. Niger) the trigger threshold is statistically indistinguishable from climatological odds. need to be cautious about using the forecast in such places.

% Max to write with Abdel's input.

\section{Conclusions and best practices}

From the abstract:  Examining United Nations operational AA forecasts in sub-Saharan Africa, we demonstrate how our method provides a better quantification of critical underlying uncertainty, enabling these AA systems to meet the state-of-the-art demanded by donors and the WMO.

% Max to draft and Dan to refine.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments
Keep acknowledgments (note correct spelling: no ``e'' between the ``g'' and
``m'') as brief as possible. In general, acknowledge only direct help in
writing or research. Financial support (e.g., grant numbers) for the work
done, for an author, or for the laboratory where the work was performed is
best acknowledged here rather than as footnotes to the title or to an
author's name. Contribution numbers (if the work has been published by the
author's institution or organization) should be included as footnotes on the title page,
not in the acknowledgments.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DATA AVAILABILITY STATEMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
%
\datastatement
The data availability statement is where authors should describe how the data underlying 
the findings within the article can be accessed and reused. Authors should attempt to 
provide unrestricted access to all data and materials underlying reported findings. 
If data access is restricted, authors must mention this in the statement.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIXES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Use \appendix if there is only one appendix.
%\appendix

% Use \appendix[A], \appendix[B], if you have multiple appendixes.
%\appendix[A]

%% Appendix title is necessary! For appendix title:
%\appendixtitle{}

%%% Appendix section numbering (note, skip \section and begin with \subsection)
% \subsection{First primary heading}

% \subsubsection{First secondary heading}

% \paragraph{First tertiary heading}

%% Important!
%\appendcaption{<appendix letter and number>}{<caption>} 
%must be used for figures and tables in appendixes, e.g.,
%
%\begin{figure}
%\noindent\includegraphics[width=19pc,angle=0]{figure01.pdf}\\
%\appendcaption{A1}{Caption here.}
%\end{figure}
%
% All appendix figures/tables should be placed in order AFTER the main figures/tables, i.e., tables, appendix tables, figures, appendix figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make your BibTeX bibliography by using these commands:
% \bibliographystyle{ametsoc2014}
% \bibliography{references}

\bibliographystyle{ametsoc2014}
\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TABLES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Enter tables at the end of the document, before figures.
%%
%
%\begin{table}[t]
%\caption{This is a sample table caption and table layout.  Enter as many tables as
%  necessary at the end of your manuscript. Table from Lorenz (1963).}\label{t1}
%\begin{center}
%\begin{tabular}{ccccrrcrc}
%\hline\hline
%$N$ & $X$ & $Y$ & $Z$\\
%\hline
% 0000 & 0000 & 0010 & 0000 \\
% 0005 & 0004 & 0012 & 0000 \\
% 0010 & 0009 & 0020 & 0000 \\
% 0015 & 0016 & 0036 & 0002 \\
% 0020 & 0030 & 0066 & 0007 \\
% 0025 & 0054 & 0115 & 0024 \\
%\hline
%\end{tabular}
%\end{center}
%\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FIGURES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Enter figures at the end of the document, after tables.
%%
%
%\begin{figure}[t]
%  \noindent\includegraphics[width=19pc,angle=0]{figure01.pdf}\\
%  \caption{Enter the caption for your figure here.  Repeat as
%  necessary for each of your figures. Figure from \protect\cite{Knutti2008}.}\label{f1}
%\end{figure}

\end{document}

### 2025-08-22 paper.tex
Some small text tweaks and updating of comments

@@
 
-\subsection{The PyCPT seasonal forecasting method}
+\subsection{The CPT seasonal forecasting method}
 
 % Max to write but could probably use Kevin and Nathan's eyes to make sure I'm not saying anything scientifically out of pocket.
-% Dan style note: "PyCPT" instead of "NextGen" 
 % Talk about heuristic rules that the country offices were using, and how they are similar conceptually / functionallty to bootstrapping
@@
 Can draw from a lot of prior IRI work here.
-Key talking points are that NextGen is a method to produce downscaled and bias-corrected seasonal forecasts using a multi-model ensemble of GCMs.
+Key talking points are that CPT is a method to produce downscaled and bias-corrected seasonal forecasts using a multi-model ensemble of GCMs.
 The models, predictand and calibration parameters can be customized to each locality and context.
@@
 Forecasts are presented as the probability of non-exceedance at a given percentile threshold (not only terciles).
+
+Talk a little bit about how the CPT method accounts for observational and model uncertainty.
 
@@
 
-\section{Sources of forecast uncertainty}
-
-% Nathan - can fold most of this into section 2b, since it isn't really a standalone section
-
-% Kevin to write w Max and Nathan as needed.
-% The basic strategy is to go through all of the different commonly discussed sources of uncertainty, and talk about how they are addressed in the NextGen approach, except for the problem of the limited historical record.
-That is what we're addressing with this paper.
-
-% note Aug 13: Can pare down the discussion of the first four sources of uncertainty, since they aren't the main point.
-
-\subsection{Uncertainty in modeling choices}
-
-Different GCMs have different assumptions about how things work. We account for this by using a multi-model ensemble in predictions.
-
-\subsection{Uncertainty in initial conditions}
-
-I'm not sure if or how to talk about this -- I'm mostly familiar with this in the context of long-term climate change projections, not seasonal forecasting.
-As I understand it, all of the GCM output that we use for seasonal forecasting is initialized with the same conditions for a given issue month.
-So maybe this is not a huge issue the way it is for CC projections, where uncertainty in initial conditions can propagate further (cf Schwarzwald and Lenssen 2022).
-But maybe I'm also missing something, so it would be good to get your input.
-
-In any case, we can talk about how we use forecasts at different lead times, with the decision rule putting greater weight on forecasts with a shorter lead time, since we are more certain about the initial conditions.
-
-\subsection{Uncertainty in accuracy of predictions}
-
-The bias correction and calibration components of NextGen are meant to account for this.
-I need to bug Andy R.
-and Azhar about this since they have promised a more thorough writeup.
-Talk about the sharpness ratio of quantifying the confidence of the forecast (this is something that decision-makers should also be thinking about, and not only whether the forecast is above or below the threshold).
-
-Note that this section is mostly focusing on how well forecasts predict observed rainfall.
-There is a second-order question about whether forecasted rainfall is related to humanitarian impacts of drought (e.g.
-hunger, loss of animals), which is of practical importance to the projects, but not really what this paper is about.
-We may want to at least mention it in passing.
-
 \subsection{Uncertainty due to the limited historical record}
@@
 
-% Nathan to write w Kevin and Max and needed
-
-% MAX TODO: Make sure we are using block randomization, with consecutive years. This could help deal with the autocorrelation problem. There is a literature we can tie this to
-
-% Abdel is particularly interested in the question of whether interannual correlation and / or trends make the calculation of confidence intervals using this method problematic.
-i need to put on my stats hat this this one.
-
-% Note that if we were really doing this properly, I think we would want to do the bootstrap sampling during the step of per-pixel model calibration (which happens in CPT), rather than doing it after the fact with areal average data as we are doing here.
-But that is a matter for a future paper.
-We want to at least mention it though I think.
-
-% kevin note: we are thinking about this as applicable to things beyond PyCPT, tho, so dont get too hung up about this
+% Nathan to write w Kevin and Max and needed. Tag in Abdel to see if this addresses his concerns about correlation
 
@@
 Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error.
+MAKE SURE WE TALK ABOUT BLOCK RANDOMIZATION HERE
 
@@
 
-% Need to be sure we are using a definition of the percentile that is consistent with what the API uses! R's may not be the same. make sure we are consistent
-
 % add table of country results, but dont get too into the weeds on country specifics
@@
 
-% tweak visual so that user focuses on the boxplot, not the clim odds
-
-% Note: The figures seem to get added to the end instead of in-line. I'm guessing that's how it is in the journal?
-
-% note that this method can acommodate compound / multimonth rules
+% note that this method can acommodate compound / multimonth rules - be sure to mention that
 

### 2025-08-22 forecast_robustness.R
New style of result visualiation

@@ -59,21 +59,40 @@ for (country in countries) {
   
   bootstrap_results <- trigger_uncertainty(admin_0_triggers,5)
   
-  plot <- ggplot(bootstrap_results,aes(y=new_threshold,x=month)) + 
-    facet_grid(cols=vars(freq)) +
-    geom_boxplot()  +
-    geom_point(aes(y=old_threshold,x=month),color='red') + 
-    geom_text(data = bootstrap_results %>% group_by(month,freq) %>% summarise_all(max), aes(y=old_threshold,x=month,label=round(old_threshold,2)),color='red',alpha=0.7,vjust=-0.5) +
-    geom_hline(aes(yintercept = as.numeric(freq)),linetype='dashed') +
-    theme_bw() +
-    ylim(c(0,50)) +
-    xlab("Issue month") +
-    ylab("Prob. non-exceedence") +
-    ggtitle(paste0(country," trigger uncertainty range"),subtitle = "AA threshold in red; clim. odds dotted line")
+  # plot <- ggplot(bootstrap_results,aes(y=new_threshold,x=month)) + 
+  #   facet_grid(cols=vars(freq)) +
+  #   geom_boxplot()  +
+  #   geom_point(aes(y=old_threshold,x=month),color='red') + 
+  #   geom_text(data = bootstrap_results %>% group_by(month,freq) %>% summarise_all(max), aes(y=old_threshold,x=month,label=round(old_threshold,2)),color='red',alpha=0.7,vjust=-0.5) +
+  #   geom_hline(aes(yintercept = as.numeric(freq)),linetype='dashed') +
+  #   theme_bw() +
+  #   ylim(c(0,50)) +
+  #   xlab("Issue month") +
+  #   ylab("Prob. non-exceedence") +
+  #   ggtitle(paste0(country," trigger uncertainty range"),subtitle = "AA threshold in red; clim. odds dotted line")
   
-  png(file=paste0(getwd(),"/figures/",country,".png"),width=650,height=400)
+  empirical <- admin_0_triggers %>% filter(month == max(month), freq == max(freq))
+  bootstrap <- bootstrap_results %>% filter(month == max(month), freq == max(freq))
+  
+  colors <- c("Cumulative Forecasts, 1983-present"="black",
+              "Empirical Trigger Threshold"="red",
+              "Bootstrapped Probability of Passing Threshold" = "blue")
+  
[…]

### 2025-08-23 paper.tex
First draft complete

@@
 % Max to draft and Dan to refine. 
+% need to intro WFP a little bit more here.
 
-Humanitarian agencies are trying to be less reactive...
-advances in the state of the science on seasonal forecasting make early action against drought feasible...
-But forecasts need to follow standards of consistency, robustness and transparency a la WMO guidelines, especially when lives are on the line...
-it is not always obvious how to pragmatically account for uncertainty under those guidelines...
-our paper advances a simple method for quantifying one type of uncertainty, which is how sensitive the trigger definition is to the limited historical record...
-this method was developed to respond to practical needs from WFP, and can be replicated elsewhere.
+Increasingly, the international community is turning to forecast-based anticipatory action to mitigate the humanitarian impacts of extreme weather events [cite].
+This shift from a reactive to a proactive model of humanitarian assistance is made possible through scientific advances in medium-term forecasting, which allow skillful forecasts of seasonal total rainfall to be made as far as 3-6 months in advance of the season, depending on the region.
+In contrast to short-term weather forecasting, seasonal forecasting is a comparatively new science, and its policy application is even newer - the UN World Food Programme has only been using seasonal forecasts for humanitarian action since the mid-2010s.
+
+This emerging field of practice demands new standards of consistency, robustness and transparency, particularly when the human costs of inaction are high.
+Accordingly, [OCHA?
+WFP?] has adopted the World Meteorological Organization (WMO) Guidelines for Seasonal Forecasting [cite], which mandate that such forecasts be rooted in process-based modeling of hydrological conditions, and use statistical methods that can accurately represent both observation- and model-based uncertainty in a probabilistic prediction.
+
+However, these guidelines do not mandate how to account for one very practical source of uncertainty: The uncertain threshold for extreme weather events due to the limited historical record.
+This type of uncertainty matters because forecast-based anticipatory action decision rules are typically based on some rare event threshold; for example, drought action may be triggered only during the driest 20 percent of years.
+In practice, forecasts often fall just on either side of this threshold, and decision makers must exercise discretion in determining whether or not to act in these situations.
+
+A more systematic way of understanding this uncertainty is needed.
+Accordingly, this paper presents a bootstrap resampling-based method for estimating the probability of mis-triggering anticipatory action for rare events.
+This method was developed to meet a practical need expressed by WFP.
+We demonstrate its application to actual seasonal forecast data used for WFP's drought anticipatory action projects, and discuss how the incorporation of our threshold uncertainty methodology addresses crucial aspects of practical decision-making that conventional metrics of predictive skill alone cannot.
+We conclude with some discussion of how this methodology could be applied more broadly.
 
@@
 
-When did the UN start using forecasts for humanitarian action?
-What about seasonal forecasts in particular?
-What obstacles (scientific and practical) have kept this from happening sooner?
-Some parallels to similar problems in the area of parametric insurance.
+According to WFP's 2024 Anticipatory Action - A Decade in Review report, "WFP expanded its Anticipatory Action operations from covering 4.1 million people in 36 countries in 2023 to over 6.2 million people in 44 countries in 2024" [cite].
+In 2024 alone, WFP's AA programs triggered over US27 million in early action programming, reaching nearly 9 million people.
+WFP's AA programming includes forecasts of drought, flood and cyclones; this paper focuses on seasonal drought forecasting as a case study, but the practical problems of extreme event forecast interpretation described herein are applicable across a variety of timescales and hazards.
+
+As a member of United Nations interagency Central Emergency Response Fund (CERF), WFP is subject to common standards for transparency, accuracy and reliable of its forecast-based AA programs.
+These standards are particularly crucial when negotiating the conditions under which AA donor funds will be disbursed - the so-called "anticipatory action plan" (AAP).
+For seasonal forecasts of drought, the technical standards are primarily determined by the WMO Seasonal Forecasting Guidelines [cite, and make sure you're naming them correctly].
+
+The primary concern of the WMO Guidelines is ensuring the seasonal forecasts are physically plausible, not overfit to limited data, and accurately represent the various sources of uncertainty in their probabilistic predictions.
+For drought forecasts, WFP predominately relies on the Climate Predictability Tool (CPT) forecasting system [cite], which was developed with the WMO guidelines in mind.
 
@@
 
-Talk about the types of seasonal forecasts we are using here, where they come from and how to interpret them.
-Can draw from a lot of prior IRI work here.
-Key talking points are that CPT is a method to produce downscaled and bias-corrected seasonal forecasts using a multi-model ensemble of GCMs.
-The models, predictand and calibration parameters can be customized to each locality and context.
-Local met services use the method and software to create purpose-built forecast models for their country or region.
-Forecasts are presented as the probability of non-exceedance at a given percentile threshold (not only terciles).
+The CPT seasonal forecasting system is based on downscaling and bias-correcting modeled precipitation conditions from global climate models (GCMs), and combining predictions from many GCMs into a probabalistic multi-model ensemble (MME).
+The models, predictand and calibration parameters can be customized to each locality and context in which the model is used [cite].
+For WFP's AA projects, local meteorological services in each country collaborate with WFP and [Columbia] to use the CPT method and software to create purpose-built forecast models for their country or region.
 
-Talk a little bit about how the CPT method accounts for observational and model uncertainty.
+In technical terms, "CPT uses the principle of cross-validation to build and validate multiple linear regression models, and to select the best model....
+CPT provides two types of pattern regression: Canonical Correlational Analysis (CCA) and Principal Components Regression (PCR)" [cite].
+In either case, CPT uses historical cross-validation to train the CCA/PCR model for each candidate GCM, and presents statistics on its locally calibrated skill.
+Based on that skill assement, the user selects the GCMs they wish to include in the MME.
+Then, "Probabilistic forecasts are created by assuming a parametric form for the forecast PDF (for example a normal bell-curve).
+The regression model provides the mean of the forecast PDF; its variance (spread) is determined from the errors of the regression model’s cross-validated hindcasts.
+Setting the forecast spread to equal the model’s error-variance ensures that the resulting probabilistic forecasts are well calibrated - i.e., that their confidence reflects their historical accuracy" [cite].
+
+The resulting forecasts are presented as the probability of non-exceedance at a given percentile threshold, which can be defined based on the scenario.
+For example, many drought anticipatory action programs use a prediction of the probability of non-exceedance below the 20th percentile of observed historical rainfall.
+However, the forecast alone cannot determine the criteria for anticipatory action (a crucial part of the AAP).
+To do so, stakeholders must exercise some judgment in determining a numeric threshold of forecast confidence for triggering action; the following section describes the typical procedure employed.
 
@@
 
-Talk about the mechanics of how we define a forecast action trigger here. Below is how I wrote it up for WFP in Ethiopia:
-
 A core component of the Anticipatory Action Plan is the forecast “trigger”, i.e., the quantitative threshold of predictive certainty above which a certain anticipatory action will be taken.
-This document describes the procedure used to determine triggers for the WFP AA project and the rationale for why it was implemented this way.
-At the end of this document, there are links to a set of live documents which provide the most up to date trigger information for each active region and woreda in the AA project.
-
 The basic rationale behind the choice of triggers is that they should be transparent, consistently applied and easily related to the expected frequency with which action will be taken.
@@
 
-Accordingly, the trigger rule used for this AAP is based on a consistent frequency of action for a given type of programming, across locations and times of year.
-For instance, in Somali Region, “Severe” level actions are associated with a 20 percent frequency of action.
+Accordingly, the trigger rule used for WFP's AA programs is based on a consistent frequency of action for a given type of programming, across locations and times of year.
+For instance, in the Somali Region of Ethiopia, “Severe” level actions are associated with a 20 percent frequency of action.
 This means that for a given location, for the forecast issued in a given month, we should expect Severe actions to trigger during the strongest 20pct of drought forecasts.
@@
 
-Talk a little about the sensitivity of this method to the choice of threshold, and what to do when the forecast is just a couple pp above or below the threshold.
-This is something that practitioners always have to deal with, and they have heuristic ways of managing it, but those can be inconsistent.
-We propose here a more systematic solution...
-segues into discussion about sources of uncertainty.
+This method is highly sensitive to the choice of numeric threshold.
+Even though the CPT methodology calibrates for observational and model uncertainty, the rare event threshold itself is subject to the inherent limitations of our comparatively short observational record (which begins in the 1980s for commonly used gridded rainfall products like CHIRPS).
+Consequently, decision-makers must exercise some judgment when a seasonal forecast is just above or just below the agreed-upon trigger threshold.
+This is typically done through heuristic means, such as adjusting the trigger threshold down by a few percentage points when observational indicators of a potential famine (such as below-average vegetation conditions leading up to the season, or assessments of food insecurity from the Integrated Phase Classification [cite]) are present.
+However, these heuristic approaches are highly subjective.
+To meet the need for a robust, transparent decision support system, a more systematic way of assessing threshold uncertainty is needed.
 
@@
 
-What we wrote for the abstract: The observational rainfall record in many regions with AA programs is limited, and interannual and interdecadal autocorrelation in the climate patterns driving seasonal rainfall further reduce the effective sample size used to calculate these thresholds \citep{martinez_seasonal_2022}.
+To pose the practical problem that decision-makers face in theoretical terms: The observational rainfall record in many regions with AA programs is limited, and interannual and interdecadal autocorrelation in the climate patterns driving seasonal rainfall further reduce the effective sample size used to calculate these thresholds \citep{martinez_seasonal_2022}.
 Thus, AA triggers are highly sensitive to the limited observational record, increasing the risk that action is not triggered when it is needed or that programs become financially unsustainable due to excessive triggers.
-In practice, forecasters are aware of this uncertainty, and ad hoc decisions on whether to trigger action are often made if seasonal forecasts are close to the threshold.
-Seeking a more formalized, statistically sound alternative to ad hoc approaches, governments and humanitarian agencies have asked for transparent and intuitive metrics for forecast trigger uncertainty.
+
+To address this problem, we need some statistical way of assessing the \textit{range} of plausible threshold values.
+Then, decision-makers can determine whether or not to exercise discretion in the decision to trigger action by assessing whether a given forecast value is above/below the threshold due to chance alone (in whih case discretion might be reasonably exercised), or whether it is statistically distinguishable from the threshold.
+
+Rare event uncertainty estimation can be parametric (such as modeling with the GEV or Gumbell distributions [cite]) or non-parametric (such as bootstrap resambling).
+Since we know there is a high degree of interannual (ex., ENSO) and interdecadal (ex.
+climate change) dependency in the historical climate record, we opt for bootstrap resampling in this case [cite, and draw out the sampling theory here a little bit more].
+Our goal is to present a simple, transparent method that can be applied to a variety of forecast types.
 
@@
 
-Below is what I've written about the method for WFP. We need to motivate why we're using the method of bootstrapping and why it's reasonable in this context first.
-
-The AA protocol is based on a binary trigger - if the drought forecast is above a given threshold of confidence, it triggers; if it is below this threshold, it does not.
-
 To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis.
@@
 Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error.
-MAKE SURE WE TALK ABOUT BLOCK RANDOMIZATION HERE
+In this case, we run 100 iterations of the re-sampling process.
+In each iteration, consecutive blocks of 5 years are randomly omitted from the historical record and then the threshold is re-computed.
+The block width of 5 was chosen here to account for the timescale of common sources of interannual correlation like ENSO, but could be adjusted by the user as they wish.
 
-In this case, we run 100 iterations of the re-sampling process. In each iteration, 5 years are randomly omitted from the historical record and then the threshold is re-computed.
-
-The box plots below depict the resulting sensitivity range in the trigger threshold for each issue month and severity level.
-The actual trigger threshold is shown in red.
-For reference, the climatological odds (the probability of non-exceedence if we were to guess from historical average conditions, without any forecast information) is shown as the dotted line.
+Figures XX shows the results of this method.
+The cumulative distribution of hindcasts is shown by the black slope.
+The empirical trigger threshold (here, the 20th percentile) is shown by the dotted red vertical line.
+The CDF of boostrapped replicates of the trigger threshold is shown by the blue slope.
+For instance, this figure tells us that based on bootstrapped resampling, there is is an estimated XX probability of the true trigger value being greater or equal to its empirical value of YY percentage points.
 
@@
 If the current forecast falls within the sensitivity range, this indicates that we should interpret the trigger result for the current season with caution.
+
+The following section demonstrates the application of this method to actual seasonal forecast data from XX of WFP's AA projects, generated by local meteorological services using the CPT software.
 
@@
 
-% note that this method can acommodate compound / multimonth rules - be sure to mention that
+% consider adding some visual indication of the 25-75 pctile spread to these figures?
 
@@
 
-- in general, the range of uncertainty tends to be a couple of percentage points in either direction.
-- forecast sharpness generally increases as the lead time decreases, but the threshold uncertainty doesn't.
-- likewise, madagascar and guatemala have the sharpest forecasts (which makes some sense; we have reason to think that ENSO teleconnections are relatively stronger there), but have as much or more threshold uncertainy than the other countries.
-- this highlights the fact that we have to think about the innate uncertainty from having a short historical record, even in places where we think the forecast model has historically performed well!
-- also in some places (ex. Niger) the trigger threshold is statistically indistinguishable from climatological odds. need to be cautious about using the forecast in such places.
+We can draw a few practical conclusions about WFP's seasonal drought forecasts from this analysis:
 
-% Max to write with Abdel's input.
+- In general, the range of uncertainty tends to be a couple of percentage points in either direction.
+
+- The forecast sharpness [explain this] generally increases as the lead time decreases, but the threshold uncertainty doesn't.
+
+- Likewise, Madagascar and Guatemala have the sharpest forecasts (unsurprisingly, as we have reason to think that ENSO teleconnections are relatively stronger there), but have as much or more threshold uncertainy than the other countries.
+
+- This highlights the fact that we have to think about the innate uncertainty from having a short historical record, even in places where we think the forecast model has historically performed well
+
+- In some places (ex. Niger) the trigger threshold is statistically indistinguishable from climatological odds. We need to be cautious about using the forecast in such places.
 
@@
 
-From the abstract:  Examining United Nations operational AA forecasts in sub-Saharan Africa, we demonstrate how our method provides a better quantification of critical underlying uncertainty, enabling these AA systems to meet the state-of-the-art demanded by donors and the WMO.
+% Max to draft and Dan to refine.
 
-% Max to draft and Dan to refine.
+% note that this method can acommodate compound / multimonth rules - be sure to mention that
+
+From the abstract, need to elaborate: Examining United Nations operational AA forecasts in sub-Saharan Africa, we demonstrate how our method provides a better quantification of critical underlying uncertainty, enabling these AA systems to meet the state-of-the-art demanded by donors and the WMO.
 

### 2025-08-26 forecast_robustness.R
Fixed bootstrapping results per Nathan input

@@ -57,28 +57,14 @@ for (country in countries) {
   
   admin_0_triggers <- compile_data_national(forecast=admin_0_rain,badyears=admin_0_rain,ndvi=NA,prevseas=NA)
   
-  bootstrap_results <- trigger_uncertainty(admin_0_triggers,5)
-  
-  # plot <- ggplot(bootstrap_results,aes(y=new_threshold,x=month)) + 
-  #   facet_grid(cols=vars(freq)) +
-  #   geom_boxplot()  +
-  #   geom_point(aes(y=old_threshold,x=month),color='red') + 
-  #   geom_text(data = bootstrap_results %>% group_by(month,freq) %>% summarise_all(max), aes(y=old_threshold,x=month,label=round(old_threshold,2)),color='red',alpha=0.7,vjust=-0.5) +
-  #   geom_hline(aes(yintercept = as.numeric(freq)),linetype='dashed') +
-  #   theme_bw() +
-  #   ylim(c(0,50)) +
-  #   xlab("Issue month") +
-  #   ylab("Prob. non-exceedence") +
-  #   ggtitle(paste0(country," trigger uncertainty range"),subtitle = "AA threshold in red; clim. odds dotted line")
-  
-  empirical <- admin_0_triggers %>% filter(month == max(month), freq == max(freq))
-  bootstrap <- bootstrap_results %>% filter(month == max(month), freq == max(freq))
+  select_data <- admin_0_triggers %>% filter(month == max(month), freq == max(freq)) ## at some point we need to update this for all freqs, months
+  bootstrap <- trigger_uncertainty_new(select_data,5) ## at some point we need to make this the default uncertainty function
   
   colors <- c("Cumulative Forecasts, 1983-present"="black",
               "Empirical Trigger Threshold"="red",
               "Bootstrapped Probability of Passing Threshold" = "blue")
   
-  plot <- ggplot(empirical,aes(x=forecast)) + stat_ecdf(aes(color="Cumulative Forecasts, 1983-present")) + 
+  plot <- ggplot(select_data,aes(x=forecast)) + stat_ecdf(aes(color="Cumulative Forecasts, 1983-present")) + 
     stat_ecdf(data=bootstrap,aes(x=new_threshold,color='Bootstrapped Probability of Passing Threshold')) +
     geom_vline(data=bootstrap,aes(xintercept = old_threshold,color='Empirical Trigger Threshold'),linetype='dashed') +
     scale_color_manual(values=colors) +

### 2025-08-26 paper.tex
latest text updates

@@
 
-Rare event uncertainty estimation can be parametric (such as modeling with the GEV or Gumbell distributions [cite]) or non-parametric (such as bootstrap resambling).
+Rare event uncertainty estimation can be parametric (such as modeling with the GEV or Gumbell distributions [cite]) or non-parametric (such as bootstrap resampling).
 Since we know there is a high degree of interannual (ex., ENSO) and interdecadal (ex.
@@
 To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis.
-This analysis is based on the method of re-estimating the trigger threshold with years randomly omitted from the historical record.
+This analysis is based on the method of re-estimating the trigger threshold with resampled draws of the historical record.
 Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error.
-In this case, we run 100 iterations of the re-sampling process.
-In each iteration, consecutive blocks of 5 years are randomly omitted from the historical record and then the threshold is re-computed.
+In this case, we run 1,000 iterations of the re-sampling process.
+In each iteration, we draw a sample of equal size to the original data comprised of randomly sampled blocks of 5 consecutive years, sampled with replacement.
 The block width of 5 was chosen here to account for the timescale of common sources of interannual correlation like ENSO, but could be adjusted by the user as they wish.
@@
 The empirical trigger threshold (here, the 20th percentile) is shown by the dotted red vertical line.
-The CDF of boostrapped replicates of the trigger threshold is shown by the blue slope.
+The CDF of bootstrapped replicates of the trigger threshold is shown by the blue slope.
 For instance, this figure tells us that based on bootstrapped resampling, there is is an estimated XX probability of the true trigger value being greater or equal to its empirical value of YY percentage points.
@@
 % fig 2: example using actual WFP data. can just be one area-month-freq. we can put the rest in an appendix.
-
-% consider adding some visual indication of the 25-75 pctile spread to these figures?
 

### 2025-08-26 forecast_robustness.R
Another figure update

@@ -62,15 +62,24 @@ for (country in countries) {
   
   colors <- c("Cumulative Forecasts, 1983-present"="black",
               "Empirical Trigger Threshold"="red",
-              "Bootstrapped Probability of Passing Threshold" = "blue")
+              "Bootstrapped Probability of Passing Threshold" = "blue",
+              "Current Forecast"="green")
   
+  current_forecast <- select_data$forecast[select_data$year == max(select_data$year)]
+  forecast_cdf <- ecdf(select_data$forecast)
+  current_pctile = forecast_cdf(current_forecast)
+  freq_shown <- max(admin_0_triggers$freq)
+                                           
   plot <- ggplot(select_data,aes(x=forecast)) + stat_ecdf(aes(color="Cumulative Forecasts, 1983-present")) + 
     stat_ecdf(data=bootstrap,aes(x=new_threshold,color='Bootstrapped Probability of Passing Threshold')) +
     geom_vline(data=bootstrap,aes(xintercept = old_threshold,color='Empirical Trigger Threshold'),linetype='dashed') +
+    geom_rect(xmin=quantile(bootstrap$new_threshold,0.10),xmax=quantile(bootstrap$new_threshold,0.90),ymin=Inf,ymax=-Inf,fill='blue',alpha=0.01) +
+    geom_point(x=current_forecast,y=current_pctile,size=3,aes(color='Current Forecast')) +
     scale_color_manual(values=colors) +
-    labs(x = "Forecast Strength", y = "Probability", color = "Legend") +
+    labs(x = "Forecast Strength", y = "Cumulative Probability", color = "Legend") +
     scale_y_continuous(labels = scales::percent) +
-    ggtitle(paste0("Forecast Trigger Uncertainty Range For ",country))
+    ggtitle(paste0("Forecast Trigger Uncertainty Range For ",country),subtitle = paste0("for ",freq_shown,"th percentile drought event")) +
+    xlim(c(0,100)) 
   
   png(file=paste0(getwd(),"/figures/",country,".png"),width=650,height=350)
   print(plot)

### 2025-09-04 paper.tex
latest text update

@@
 To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis.
-This analysis is based on the method of re-estimating the trigger threshold with resampled draws of the historical record.
-Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error.
+This analysis is based on the method of re-estimating the trigger threshold with resampled draws of the historical record, known as bootstrappaing [cite].
+Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error of a limited historical record.
 In this case, we run 1,000 iterations of the re-sampling process.
 In each iteration, we draw a sample of equal size to the original data comprised of randomly sampled blocks of 5 consecutive years, sampled with replacement.
-The block width of 5 was chosen here to account for the timescale of common sources of interannual correlation like ENSO, but could be adjusted by the user as they wish.
+The block width of 5 was chosen here to account for the timescale of common sources of interannual correlation like ENSO.
 
@@
 
-If the current forecast is outside of the sensitivity range, this indicates that we have statistically robust evidence that it is above or below the threshold.
-If the current forecast falls within the sensitivity range, this indicates that we should interpret the trigger result for the current season with caution.
+If the current forecast is outside of the sensitivity range, this indicates that we have statistically robust evidence that the forecasted conditions are above or below the threshold.
+If the current forecast falls within the sensitivity range, this indicates that we should interpret the trigger result for the current season with caution as the appearance of trigger-worth conditions could be occurring just by chance.
 

### 2025-09-04 forecast_robustness_madagascar.R
Added data and bootstrap results as file outputs; organized folders

@@ -18,11 +18,11 @@ month_names <- c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","no
 
 # country / project
 
-countries <- c("madagascar","ethiopia","niger","guatemala-jja","djibouti-ond")
+countries <- c("madagascar")
 
 # freqs (0-100)
 
-freqs <- c(20,30)
+freqs <- c(15)
 
 # season (1 or 2)
 
@@ -30,14 +30,14 @@ season <- 'season1'
 
 # months (numbered 0-11)
 
-months_list <- list(c(8, 9, 10),c(11, 0, 1),c(3, 4, 5),c(2, 3, 4),c(6, 7, 8)) 
+months_list <- list(c(8, 9, 10)) 
 months_df <- data.frame("month" = c(0:11),"name" = month_names)
 prev_months_df <- data.frame("month" = c(2:11,0,1),"name" = month_names)
 
 # define dataset names
 
-forecast_list <- c("pnep-v3","pnep-v5","prcp-v5","pnep-v6","pnep-v2")
-bad_years_list <- c("bad-years-v3","bad-years","bad-years-v3","bad-years","bad-years-rank-v2") 
+forecast_list <- c("pnep-v3")
+bad_years_list <- c("enacts-precip-all") 
 
 # server (e.g. production or test)
 
@@ -46,8 +46,6 @@ server <- "https://iridl.ldeo.columbia.edu/fbfmaproom2/"
 i <- 1
 for (country in countries) {
 
-  print(paste0("Running for ",country))
-  
   ## get key file
   
   admin_0_key <- get_key(server,country,0)
@@ -59,39 +57,38 @@ for (country in countries) {
   
   admin_0_triggers <- compile_data_national(forecast=admin_0_rain,badyears=admin_0_rain,ndvi=NA,prevseas=NA)
   
-  select_data <- admin_0_triggers %>% filter(month == max(month), freq == max(freq)) 
-  
-  write.csv(select_data,paste0(getwd(),"/1. data/",country,".csv"),row.names = FALSE)
-  
+  # select_data <- admin_0_triggers %>% filter(month == max(month), freq == max(freq)) 
+  select_data <- admin_0_triggers
   bootstrap <- trigger_uncertainty(select_data,5) 
   
   colors <- c("Cumulative Forecasts, 1983-present"="black",
               "Empirical Trigger Threshold"="red",
               "Bootstrapped Probability of Passing Threshold" = "blue",
-              "Current Forecast"="green")
[…]

### 2025-09-20 paper.tex
latest text update, w/ nathan's edits

@@
 To test how sensitive the protocol is to the exact threshold values chosen, this section presents an uncertainty analysis.
-This analysis is based on the method of re-estimating the trigger threshold with resampled draws of the historical record, known as bootstrappaing [cite].
+This analysis is based on the method of re-estimating the trigger threshold with resampled draws of the historical record, known as bootstrapping [cite].
 Repeating this procedure many times gives us an estimated distribution of how much the threshold might be expected to vary due to sampling error of a limited historical record.
@@
 % fig 2: example using actual WFP data. can just be one area-month-freq. we can put the rest in an appendix.
+
+% take a look at Kevin's github for some references of how to code this
+
+% discuss how uncertainty increases as frequency increases
 

### 2025-09-20 forecast_robustness.R
Latest figure updates

@@ -7,6 +7,10 @@
 
 source("~/iri-fist/aa_dashboard_common_code/aa_functions.R")
 setwd("~/forecast_uncertainty")
+library(ggpubr)
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(cowplot)
 
 options(timeout=1000000000000)
 
@@ -17,8 +21,9 @@ month_names <- c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","no
 ## set what data you'd like to download
 
 # country / project
+# replace guatemala with lesotho, add another west african country 
 
-countries <- c("madagascar","ethiopia","niger","guatemala-jja","djibouti-ond")
+countries <- c("madagascar","ethiopia","niger","guatemala-jja","djibouti-ond") # make display names nicer
 
 # freqs (0-100)
 
@@ -43,6 +48,8 @@ bad_years_list <- c("bad-years-v3","bad-years","bad-years-v3","bad-years","bad-y
 
 server <- "https://iridl.ldeo.columbia.edu/fbfmaproom2/"
 
+plots <- list()
+ranges <- list()
 i <- 1
 for (country in countries) {
 
@@ -65,7 +72,7 @@ for (country in countries) {
   
   bootstrap <- trigger_uncertainty(select_data,5) 
   
-  colors <- c("Cumulative Forecasts, 1983-present"="black",
+  colors <- c("Forecasts, 1983-present"="black",
               "Empirical Trigger Threshold"="red",
               "Bootstrapped Probability of Passing Threshold" = "blue",
               "Current Forecast"="green")
@@ -75,26 +82,175 @@ for (country in countries) {
   current_pctile = forecast_cdf(current_forecast)
   freq_shown <- max(admin_0_triggers$freq)
                                            
-  plot <- ggplot(select_data,aes(x=forecast)) + stat_ecdf(aes(color="Cumulative Forecasts, 1983-present")) + 
+  plots[[i]] <- ggplot(select_data,aes(x=forecast)) + stat_ecdf(aes(color="Forecasts, 1983-present")) + 
     stat_ecdf(data=bootstrap,aes(x=new_threshold,color='Bootstrapped Probability of Passing Threshold')) +
     geom_vline(data=bootstrap,aes(xintercept = old_threshold,color='Empirical Trigger Threshold'),linetype='dashed') +
[…]
